---
title: "Himalayan Expeditions: Analyzing Success and Risk in the World's Highest Mountains"
author: "Paolo Nicolet, Alessandro Mecchia"
output: html_document
---

# Final project Data Projects & Hackathon 2

Dataset: <https://github.com/rfordatascience/tidytuesday/tree/main/data/2025/2025-01-21>

## Import data and tidy conversion

```{r}
library(tidyverse)
library(foreign)
library(janitor)
library(httr)
library(glue)
library(zip)
library(ggplot2)
library(infer)
library(tidymodels)
```

```{r}
################################################################################
### Cleaning data from the Himalayan Database ##################################
################################################################################

###_____________________________________________________________________________
### Utilize httr and foreign to gain access to the file and import
###_____________________________________________________________________________

# Step 1: Set up the URL and file paths
url <- "https://www.himalayandatabase.com/downloads/Himalayan%20Database.zip"
zip_file <- tempfile(fileext = ".zip")
extracted_dir <- tempfile()

# Step 2: Download the ZIP file
httr::GET(url, httr::write_disk(zip_file, overwrite = TRUE))

# Step 3: Extract the ZIP file
zip::unzip(zip_file, exdir = extracted_dir)

# Test to confirm the extracted contents
list.files(extracted_dir, recursive = TRUE)

# Step 4: Locate the peaks.DBF and exped.DBP files
dbf_file_peaks <- file.path(extracted_dir, "Himalayan Database", "HIMDATA", "peaks.DBF")
dbf_file_exped <- file.path(extracted_dir, "Himalayan Database", "HIMDATA", "exped.DBF")
print(dbf_file_peaks) # Debugging: Verify the file path
print(dbf_file_exped) # Debugging: Verify the file path

# Step 5: Read the .DBF files using the `foreign` package
suppressWarnings({
  peaks_data <- foreign::read.dbf(dbf_file_peaks, as.is = F)
  exped_data <- foreign::read.dbf(dbf_file_exped, as.is = F)
})
# Step 6: Convert the data to a tidy tibble
peaks_temp <- tibble::as_tibble(peaks_data)
exped_temp <- tibble::as_tibble(exped_data)

# Preview the data
# originally there but we don't need that
# glimpse(peaks_temp)
# glimpse(exped_temp)

# Clean up temporary files (optional)
unlink(c(zip_file, extracted_dir), recursive = TRUE)
```

```{r}
###_____________________________________________________________________________
### Add columns that recode some variables that are integers into character
### strings that describe their subject in the peaks dataset
###_____________________________________________________________________________

peaks_tidy <- peaks_temp |> 
  dplyr::mutate(HIMAL_FACTOR = case_when(
    HIMAL == 0  ~ "Unclassified",
    HIMAL == 1  ~ "Annapurna",
    HIMAL == 2  ~ "Api/Byas Risi/Guras",
    HIMAL == 3  ~ "Damodar",
    HIMAL == 4  ~ "Dhaulagiri",
    HIMAL == 5  ~ "Ganesh/Shringi",
    HIMAL == 6  ~ "Janak/Ohmi Kangri",
    HIMAL == 7  ~ "Jongsang",
    HIMAL == 8  ~ "Jugal",
    HIMAL == 9  ~ "Kangchenjunga/Simhalila",
    HIMAL == 10 ~ "Kanjiroba",
    HIMAL == 11 ~ "Kanti/Palchung",
    HIMAL == 12 ~ "Khumbu",
    HIMAL == 13 ~ "Langtang",
    HIMAL == 14 ~ "Makalu",
    HIMAL == 15 ~ "Manaslu/Mansiri",
    HIMAL == 16 ~ "Mukut/Mustang",
    HIMAL == 17 ~ "Nalakankar/Chandi/Changla",
    HIMAL == 18 ~ "Peri",
    HIMAL == 19 ~ "Rolwaling",
    HIMAL == 20 ~ "Saipal",
    TRUE ~ "Unknown" # Default case if a value doesn't match
  ),
  .after = HIMAL
  ) |> 
  mutate(REGION_FACTOR = case_when(
    REGION == 0 ~ "Unclassified",
    REGION == 1 ~ "Kangchenjunga-Janak",
    REGION == 2 ~ "Khumbu-Rolwaling-Makalu",
    REGION == 3 ~ "Langtang-Jugal",
    REGION == 4 ~ "Manaslu-Ganesh",
    REGION == 5 ~ "Annapurna-Damodar-Peri",
    REGION == 6 ~ "Dhaulagiri-Mukut",
    REGION == 7 ~ "Kanjiroba-Far West",
    TRUE        ~ "Unknown" # Default case if a value doesn't match
  ),
  .after = REGION
  ) |> 
  mutate(PHOST_FACTOR = case_when(
    PHOST == 0 ~ "Unclassified",
    PHOST == 1 ~ "Nepal only",
    PHOST == 2 ~ "China only",
    PHOST == 3 ~ "India only",
    PHOST == 4 ~ "Nepal & China",
    PHOST == 5 ~ "Nepal & India",
    PHOST == 6 ~ "Nepal, China & India",
    TRUE       ~ "Unknown" # Default case if a value doesn't match
  ),
  .after = PHOST
  ) |> 
  mutate(PSTATUS_FACTOR = case_when(
    PSTATUS == 0 ~ "Unknown",
    PSTATUS == 1 ~ "Unclimbed",
    PSTATUS == 2 ~ "Climbed",
    TRUE         ~ "Invalid" # Default case if a value doesn't match
  ),
  .after = PSTATUS
  ) |> 
  mutate(across(c(HIMAL_FACTOR, PHOST_FACTOR, PSTATUS_FACTOR), ~ factor(.)))

```

```{r}
exped_tidy <- exped_temp |> 
  dplyr::mutate(YEAR = as.numeric(trimws(YEAR))) |> 
  dplyr::filter(YEAR >= 2000) |> 
  dplyr::mutate(SEASON_FACTOR = case_when(
    SEASON == 0 ~ "Unknown",
    SEASON == 1 ~ "Spring",
    SEASON == 2 ~ "Summer",
    SEASON == 3 ~ "Autumn",
    SEASON == 4 ~ "Winter",
    TRUE ~ NA_character_
  ), .after = SEASON) |> 
  dplyr::mutate(HOST_FACTOR = case_when(
    HOST == 0 ~ "Unknown",
    HOST == 1 ~ "Nepal",
    HOST == 2 ~ "China",
    HOST == 3 ~ "India",
    TRUE ~ NA_character_
  ), .after = HOST) |> 
  dplyr::mutate(TERMREASON_FACTOR = case_when(
    TERMREASON == 0 ~ "Unknown",
    TERMREASON == 1 ~ "Success (main peak)",
    TERMREASON == 2 ~ "Success (subpeak, foresummit)",
    TERMREASON == 3 ~ "Success (claimed)",
    TERMREASON == 4 ~ "Bad weather (storms, high winds)",
    TERMREASON == 5 ~ "Bad conditions (deep snow, avalanching, falling ice, or rock)",
    TERMREASON == 6 ~ "Accident (death or serious injury)",
    TERMREASON == 7 ~ "Illness, AMS, exhaustion, or frostbite",
    TERMREASON == 8 ~ "Lack (or loss) of supplies, support or equipment",
    TERMREASON == 9 ~ "Lack of time",
    TERMREASON == 10 ~ "Route technically too difficult, lack of experience, strength, or motivation",
    TERMREASON == 11 ~ "Did not reach base camp",
    TERMREASON == 12 ~ "Did not attempt climb",
    TERMREASON == 13 ~ "Attempt rumored",
    TERMREASON == 14 ~ "Other"
  ), .after = TERMREASON)

```

```{r}
dim(exped_temp)
dim(exped_tidy)
```

```{r}
view(peaks_tidy)
view(exped_tidy)
view(exped_temp)
```

```{r}
ggplot(exped_temp, aes(x = YEAR)) +
  geom_bar() +
  scale_x_discrete(breaks = levels(exped_temp$YEAR)[as.numeric(levels(exped_temp$YEAR)) %% 10 == 0]) +
  labs(title = "Number of Expeditions per Year",
       x = "Year",
       y = "Number of Expeditions") +
  theme_minimal()
```

## Data Cleaning

### Handling NaN

At the moment the data contain a lot of NaN, I want to check If exist some columns that are principaly composed of missing values, if exists, we will remove them

```{r}
# 1. Compute % missing per column
miss_pct <- exped_tidy %>% 
  summarise_all(~ mean(is.na(.)) * 100) %>% 
  pivot_longer(everything(), names_to = "variable", values_to = "pct_missing")

# 2. Filter for columns with > 0% missing values
miss_pct_filtered <- miss_pct %>% 
  filter(pct_missing > 0)

# 3. Plot
miss_pct_filtered %>% 
  ggplot(aes(x = pct_missing, y = fct_reorder(variable, pct_missing))) +
  geom_col() +
  labs(
    x = "% missing",
    y = NULL,
    title = "Missing values per variable in exped_tidy (only > 0%)"
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, .05)))


```

```{r}
#round(colSums(is.na(peaks_tidy)) / nrow(peaks_tidy) * 100, 2)

```

```{r}
# 1. Compute % missing per column
miss_pct <- peaks_tidy %>% 
  summarise_all(~ mean(is.na(.)) * 100) %>% 
  pivot_longer(everything(), names_to = "variable", values_to = "pct_missing")

# 2. Filter for columns with > 0% missing values
miss_pct_filtered <- miss_pct %>% 
  filter(pct_missing > 0)

# 3. Plot
miss_pct_filtered %>% 
  ggplot(aes(x = pct_missing, y = fct_reorder(variable, pct_missing))) +
  geom_col() +
  labs(
    x = "% missing",
    y = NULL,
    title = "Missing values per variable in peaks_tidy (only > 0%)"
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, .05)))
```

```{r}
miss_pct_filtered
```

| Feature        | Class     | Drop | Motivation if drop                 |
|----------------|-----------|------|------------------------------------|
| PEAKID         | integer   | No   |                                    |
| PKNAME         | integer   | No   |                                    |
| PKNAME2        | integer   | Yes  | 46.5% of Missing Values            |
| LOCATION       | integer   | No   |                                    |
| HEIGHTM        | integer   | No   |                                    |
| HEIGHTF        | integer   | Yes  | We use meter, not feet             |
| HIMAL          | integer   | No   |                                    |
| HIMAL_FACTOR   | integer   | Yes  | Already encoded                    |
| REGION         | integer   | No   |                                    |
| REGION_FACTOR  | character | Yes  | Already encoded                    |
| OPEN           | logical   | Yes  | Is not important for out reasearch |
| UNLISTED       | logical   | Yes  |                                    |
| TREKKING       | logical   | Yes  |                                    |
| TREKYEAR       | integer   | Yes  | 93.9% of Missing Values            |
| RESTRICT       | logical   | Yes  | 42.7 of Missing Values             |
| PHOST          | integer   | No   |                                    |
| PHOST_FACTOR   | character | Yes  | Already encoded                    |
| PSTATUS        | integer   | No   |                                    |
| PSTATUS_FACTOR | character | Yes  | Already encoded                    |
| PEAKMEMO       | character | Yes  |                                    |
| PYEAR          | integer   | Yes  |                                    |
| PSEASON        | character | Yes  |                                    |
| PEXPID         | integer   | Yes  |                                    |
| PSMTDATE       | character | Yes  |                                    |
| PCOUNTRY       | character | Yes  |                                    |
| PSUMMITERS     | integer   | Yes  |                                    |
| PSMTNOTE       | character | Yes  |                                    |
| REFERMEMO      | character | Yes  |                                    |
| PHOTOMEMO      | character | Yes  |                                    |

```{r}
peaks_tidy_clean <- peaks_tidy %>% 
  select(-c(
    PKNAME2,
    HIMAL_FACTOR,
    TREKYEAR,
    RESTRICT,
    PEAKMEMO,
    PSEASON,
    PEXPID,
    PSMTDATE,
    PCOUNTRY,
    PSUMMITERS,
    PSMTNOTE,
    REFERMEMO,
    PHOTOMEMO
  ))

exped_tidy_clean <- exped_tidy %>%
  select(-c(
    ROUTE2,
    ROUTE3,
    ROUTE4,
    ASCENT1,
    ASCENT2,
    ASCENT3,
    ASCENT4,
    COUNTRIES,
    APPROACH,
    SKI,
    PARAPENTE,
    TERMNOTE,
    OTHERSMTS,
    CAMPSITES,
    ROUTEMEMO,
    ACHIEVMENT,
    PRIMMEM,
    PRIMID
  ))
```

```{r}

# 1. Compute % missing per column
miss_pct <- exped_tidy_clean %>% 
  summarise_all(~ mean(is.na(.)) * 100) %>% 
  pivot_longer(everything(), names_to = "variable", values_to = "pct_missing")

# 2. Filter for columns with > 0% missing values
miss_pct_filtered <- miss_pct %>% 
  filter(pct_missing > 0)

# 3. Plot
miss_pct_filtered %>% 
  ggplot(aes(x = pct_missing, y = fct_reorder(variable, pct_missing))) +
  geom_col() +
  labs(
    x = "% missing",
    y = NULL,
    title = "Missing values per variable in exped_tidy_clean (only > 0%)"
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, .05)))

```

```{r}
ggplot(peaks_tidy, aes(x = HIMAL_FACTOR)) +
  geom_bar() +
  coord_flip() +
  labs(title = "Distribution of Peaks by HIMAL_FACTOR",
       x = "HIMAL Factor",
       y = "Count") +
  theme_minimal()
```

```{r}
annapurna_peaks <- peaks_tidy %>%
  filter(HIMAL_FACTOR == "Annapurna")

print(annapurna_peaks)
```

```{r}
nan_percent <- mean(is.na(exped_tidy$ACCIDENTS)) * 100
cat("Percentage of NA values in ACCIDENT column:", round(nan_percent, 2), "%\n")
```

```{r}
view(exped_tidy)
view(exped_temp)
```

```{r}
# Check for duplicated values in ASCENT1
duplicated_ascent1 <- exped_tidy %>%
  filter(!is.na(ASCENT1)) %>%     # remove NAs
  group_by(ASCENT1) %>%
  filter(n() > 1) %>%
  arrange(ASCENT1)

# View the duplicated entries
print(duplicated_ascent1)
```

### Supplemental Oxygen: How does the use of oxygen affect the likelihood of success?

```{r}
exped_oxygen <- exped_tidy_clean %>%
  mutate(SUCCESS = SUCCESS1 | SUCCESS2 | SUCCESS3 | SUCCESS4) %>%
  filter(!is.na(O2USED) & !is.na(SUCCESS))

```

```{r}
ggplot(exped_oxygen, aes(x = O2USED, fill = SUCCESS)) +
  geom_bar(position = "fill") +
  labs(y = "Proportion", title = "Success Rate by Oxygen Use")

```

```{r}
succes <- exped_oxygen %>%
  filter(SUCCESS)

print(succes)
```

```{r}
exped_oxygen %>%
  specify(SUCCESS ~ O2USED, SUCCESS = "TRUE") %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "diff in props", order = c("TRUE", "FALSE")) %>%
  visualize()

```

```{r}
log_spec <- logistic_reg() %>% set_engine("glm")      S
log_fit <- log_spec %>% fit(success ~ O2USED, data = exped_oxygenS)

```

```{r}
obs_stat <- exped_oxygen %>%
  specify(success ~ O2USED, success = "TRUE") %>%
  calculate(stat = "diff in props", order = c("TRUE", "FALSE"))

null_dist <- exped_oxygen %>%
  specify(success ~ O2USED, success = "TRUE") %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "diff in props", order = c("TRUE", "FALSE"))

null_dist %>%
  visualize() +
  shade_p_value(obs_stat, direction = "greater")

```

```{r}
library(tidymodels)

log_spec <- logistic_reg() %>%
  set_engine("glm")

log_fit <- log_spec %>%
  fit(success ~ O2USED, data = exped_oxygen)

summary(log_fit$fit)

```

## regression on height to the summit

```{r}
# Count and print the number of HIGHPOINT = NaN and HIGHPOINT = 0
num_nan <- exped_tidy_clean %>%
  left_join(peaks_tidy_clean, by = "PEAKID") %>%
  filter(is.na(HIGHPOINT)) %>%
  nrow()

num_zero <- exped_tidy_clean %>%
  left_join(peaks_tidy_clean, by = "PEAKID") %>%
  filter(HIGHPOINT == 0) %>%
  nrow()

cat("Number of cases with HIGHPOINT = NaN:", num_nan, "\n")
cat("Number of cases with HIGHPOINT = 0:", num_zero, "\n")
```

```{r}
library(tidymodels)
# Merge and calculate height_remaining
exped_full <- exped_tidy_clean %>%
  left_join(peaks_tidy_clean, by = "PEAKID") %>%
  filter(HIGHPOINT != 0) %>%
  mutate(height_remaining = HEIGHTM - HIGHPOINT)

# Optional: View the result
exped_full %>%
  select(height_remaining, SUCCESS1)



```

```{r}
# 1. Create a train-test split
set.seed(1)  # For reproducibility
split <- initial_split(exped_full, prop = 0.7)
train_data <- training(split)
test_data  <- testing(split)

# Remove HIGHPOINT and Height
#train_data_filtered <- train_data |> select(-HIGHPOINT, -HEIGHTM)


# 2. Define a linear regression model using parsnip
lin_mod <- linear_reg() |> set_engine("lm")

# 3. Fit the model
# Let's say you want to predict height_remaining using HIGHPOINT and maybe other features like year, season, etc.
lm_fit_all <- lin_mod |>
  fit(height_remaining ~ SEASON_FACTOR + YEAR + PEAKID, data = train_data)

# 4. See model summary
lm_fit |> extract_fit_engine() |> summary()

# 5. Check assumptions
library(performance)
lm_fit |> extract_fit_engine() |> check_model()

# 6. Predict on test data
preds <- predict(lm_fit, new_data = test_data)
head(preds)

# Or use augment to get residuals and fitted values
augmented <- augment(lm_fit, new_data = test_data)
head(augmented)

```

```{r}
# Get factor level counts for training and testing data
train_levels <- sapply(train_data_filtered, function(x) {
  if (is.factor(x)) length(unique(x)) else NA
}) |> na.omit()

test_levels <- sapply(test_data, function(x) {
  if (is.factor(x)) length(unique(x)) else NA
}) |> na.omit()

# Combine into a comparison table
level_comparison <- data.frame(
  Variable = names(train_levels),
  Train_Levels = as.integer(train_levels),
  Test_Levels = as.integer(test_levels[names(train_levels)]),
  row.names = NULL
)

# Show result
level_comparison

```

```{r}
head(exped_full)
```
