---
title: "R Notebook"
output: html_notebook
---

# Final project Data Projects & Hackathon 2

Dataset: <https://github.com/rfordatascience/tidytuesday/tree/main/data/2025/2025-01-21>

## Introduction and problem setting

This project is based on the “Himalayan Database”, a detailed collection of all documented expeditions to the Nepalese Himalayas since 1905. In particular, we will analyse data on expeditions from 2010 onwards, in order to understand the factors influencing the success or failure of recent mountaineering missions.

The main objective is to explore what factors influence the outcome of an expedition. Some key questions:

-   Which seasons have the highest probability of success?

-   Has the reason for ending an expedition changed in recent years?

-   Are there differences between expeditions led from Nepal, China or India?

-   Are there countries that are more present and more successful?

-   How much does the use of supplementary oxygen influence success?

## Import data and cleaning

```{r}
################################################################################
### Cleaning data from the Himalayan Database ##################################
################################################################################


library(tidyverse)
library(foreign)
library(httr)
library(zip)

Sys.setlocale("LC_ALL", "C")  # resolve encoding problem

# Step 1: Setup URL 
url <- "https://www.himalayandatabase.com/downloads/Himalayan%20Database.zip"
zip_file <- tempfile(fileext = ".zip")
extracted_dir <- tempfile()

# Step 2: download ZIP and exctract
GET(url, write_disk(zip_file, overwrite = TRUE))
unzip(zip_file, exdir = extracted_dir)

dbf_file_peaks <- file.path(extracted_dir, "Himalayan Database", "HIMDATA", "peaks.DBF")
dbf_file_exped <- file.path(extracted_dir, "Himalayan Database", "HIMDATA", "exped.DBF")

# Step 5: read DBF and ignore warnings (already verified it)
peaks_data <- suppressWarnings(read.dbf(dbf_file_peaks, as.is = TRUE))
exped_data <- suppressWarnings(read.dbf(dbf_file_exped, as.is = TRUE))

# Step 6: convert into Tibble
peaks_temp <- as_tibble(peaks_data)
exped_temp <- as_tibble(exped_data)

unlink(c(zip_file, extracted_dir), recursive = TRUE)
```

```{r}
###_____________________________________________________________________________
### Add columns that recode some variables that are integers into character
### strings that describe their subject in the peaks dataset
###_____________________________________________________________________________

peaks_tidy <- peaks_temp |> 
  dplyr::mutate(HIMAL_FACTOR = case_when(
    HIMAL == 0  ~ "Unclassified",
    HIMAL == 1  ~ "Annapurna",
    HIMAL == 2  ~ "Api/Byas Risi/Guras",
    HIMAL == 3  ~ "Damodar",
    HIMAL == 4  ~ "Dhaulagiri",
    HIMAL == 5  ~ "Ganesh/Shringi",
    HIMAL == 6  ~ "Janak/Ohmi Kangri",
    HIMAL == 7  ~ "Jongsang",
    HIMAL == 8  ~ "Jugal",
    HIMAL == 9  ~ "Kangchenjunga/Simhalila",
    HIMAL == 10 ~ "Kanjiroba",
    HIMAL == 11 ~ "Kanti/Palchung",
    HIMAL == 12 ~ "Khumbu",
    HIMAL == 13 ~ "Langtang",
    HIMAL == 14 ~ "Makalu",
    HIMAL == 15 ~ "Manaslu/Mansiri",
    HIMAL == 16 ~ "Mukut/Mustang",
    HIMAL == 17 ~ "Nalakankar/Chandi/Changla",
    HIMAL == 18 ~ "Peri",
    HIMAL == 19 ~ "Rolwaling",
    HIMAL == 20 ~ "Saipal",
    TRUE ~ "Unknown" # Default case if a value doesn't match
  ),
  .after = HIMAL
  ) |> 
  mutate(REGION_FACTOR = case_when(
    REGION == 0 ~ "Unclassified",
    REGION == 1 ~ "Kangchenjunga-Janak",
    REGION == 2 ~ "Khumbu-Rolwaling-Makalu",
    REGION == 3 ~ "Langtang-Jugal",
    REGION == 4 ~ "Manaslu-Ganesh",
    REGION == 5 ~ "Annapurna-Damodar-Peri",
    REGION == 6 ~ "Dhaulagiri-Mukut",
    REGION == 7 ~ "Kanjiroba-Far West",
    TRUE        ~ "Unknown" # Default case if a value doesn't match
  ),
  .after = REGION
  ) |> 
  mutate(PHOST_FACTOR = case_when(
    PHOST == 0 ~ "Unclassified",
    PHOST == 1 ~ "Nepal only",
    PHOST == 2 ~ "China only",
    PHOST == 3 ~ "India only",
    PHOST == 4 ~ "Nepal & China",
    PHOST == 5 ~ "Nepal & India",
    PHOST == 6 ~ "Nepal, China & India",
    TRUE       ~ "Unknown" # Default case if a value doesn't match
  ),
  .after = PHOST
  ) |> 
  mutate(PSTATUS_FACTOR = case_when(
    PSTATUS == 0 ~ "Unknown",
    PSTATUS == 1 ~ "Unclimbed",
    PSTATUS == 2 ~ "Climbed",
    TRUE         ~ "Invalid" # Default case if a value doesn't match
  ),
  .after = PSTATUS
  ) |> 
  mutate(across(c(HIMAL_FACTOR, PHOST_FACTOR, PSTATUS_FACTOR), ~ factor(.)))

```

```{r}
exped_tidy <- exped_temp |> 
  dplyr::mutate(YEAR = as.numeric(trimws(YEAR))) |> 
  dplyr::filter(YEAR >= 2010) |> 
  dplyr::mutate(SEASON_FACTOR = case_when(
    SEASON == 0 ~ "Unknown",
    SEASON == 1 ~ "Spring",
    SEASON == 2 ~ "Summer",
    SEASON == 3 ~ "Autumn",
    SEASON == 4 ~ "Winter",
    TRUE ~ NA_character_
  ), .after = SEASON) |> 
  dplyr::mutate(HOST_FACTOR = case_when(
    HOST == 0 ~ "Unknown",
    HOST == 1 ~ "Nepal",
    HOST == 2 ~ "China",
    HOST == 3 ~ "India",
    TRUE ~ NA_character_
  ), .after = HOST) |> 
  dplyr::mutate(TERMREASON_FACTOR = case_when(
    TERMREASON == 0 ~ "Unknown",
    TERMREASON == 1 ~ "Success (main peak)",
    TERMREASON == 2 ~ "Success (subpeak, foresummit)",
    TERMREASON == 3 ~ "Success (claimed)",
    TERMREASON == 4 ~ "Bad weather (storms, high winds)",
    TERMREASON == 5 ~ "Bad conditions (deep snow, avalanching, falling ice, or rock)",
    TERMREASON == 6 ~ "Accident (death or serious injury)",
    TERMREASON == 7 ~ "Illness, AMS, exhaustion, or frostbite",
    TERMREASON == 8 ~ "Lack (or loss) of supplies, support or equipment",
    TERMREASON == 9 ~ "Lack of time",
    TERMREASON == 10 ~ "Route technically too difficult, lack of experience, strength, or motivation",
    TERMREASON == 11 ~ "Did not reach base camp",
    TERMREASON == 12 ~ "Did not attempt climb",
    TERMREASON == 13 ~ "Attempt rumored",
    TERMREASON == 14 ~ "Other"
  ), .after = TERMREASON)

```

```{r}
# not clean dataset
dim(exped_temp)
# clean dataset
dim(exped_tidy)
```

## Data Exploration (EDA)

Information about the dataset

```{r}
glimpse(exped_tidy)
```

```{r}
# missing values for each column
missing_counts <- colSums(is.na(exped_tidy))
missing_counts[missing_counts > 0]

```

```{r}
# number of expedition by year
library(ggplot2)
ggplot(exped_tidy, aes(x = YEAR)) +
  geom_bar() +
  labs(title = "Number of expedition by year", x = "year", y = "n")
```

```{r}
ggplot(exped_tidy, aes(x = SEASON_FACTOR)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Distribution of expeditions per season", x = "Season", y = "Count")

```

```{r}
# Distribution by host
exped_tidy |> count(HOST_FACTOR) |> ggplot(aes(x = HOST_FACTOR, y = n)) +
  geom_bar(stat = "identity") + labs(title = "Distribution by host", x = "Host", y = "expedition")
```

```{r}
# expedition duration days
ggplot(exped_tidy, aes(x = TOTDAYS)) + 
  geom_histogram(bins = 30) + 
  labs(title = "Distribution of total days", x = "days", y = "freq")

# totals member
ggplot(exped_tidy, aes(x = TOTMEMBERS)) + 
  geom_histogram(bins = 30) + 
  labs(title = "Distribution of number of members", x = "tot members", y = "freq")
```

```{r}
# Success per season
exped_tidy |> 
 mutate(success = ifelse(SUCCESS1, "Success", "Failure")) |> 
  count(SEASON_FACTOR, success) |> 
 ggplot(aes(x = SEASON_FACTOR, y = n, fill = success)) +
 geom_bar(stat = "identity", position = "dodge") +
 labs(title = "Successes and Failures by Season", x = "Season", y = "Number", fill = "Outcome")
```

```{r}
# Oxygen and success
exped_tidy |> 
 mutate(O2_USED = ifelse(O2USED, "O2 Used", "No O2")) |> 
 mutate(success = ifelse(SUCCESS1, "Success", "Failure")) |> 
  count(O2_USED, success) |> 
  ggplot(aes(x = O2_USED, y = n, fill = success)) +
 geom_bar(stat = "identity", position = "dodge") +
 labs(title = "Success in relation to Oxygen", x = "Oxygen", y = "Number", fill = "Result")

```

```{r}
library(GGally)
exped_tidy_num <- exped_tidy |> select(TOTDAYS, SMTDAYS, TOTMEMBERS, SMTMEMBERS, HIGHPOINT)
GGally::ggpairs(exped_tidy_num)

```

```{r}
# Boxplot giorni
ggplot(exped_tidy, aes(y = TOTDAYS)) + 
  geom_boxplot() + 
  labs(title = "Outliers in the total expedition days", y = "tot days")

```

```{r}

# outliers in members
exped_tidy |> filter(TOTMEMBERS > 30) |> select(EXPID, TOTMEMBERS)
```

```         
```

## Modeling

## Conclusion
