---
title: "R Notebook"
output: html_notebook
---

# Final project Data Projects & Hackathon 2

Dataset: <https://github.com/rfordatascience/tidytuesday/tree/main/data/2025/2025-01-21>

## Introduction and problem setting

This project is based on the “Himalayan Database”, a detailed collection of all documented expeditions to the Nepalese Himalayas since 1905. In particular, we will analyse data on expeditions from 2010 onwards, in order to understand the factors influencing the success or failure of recent mountaineering missions.

The main objective is to explore what factors influence the outcome of an expedition. Some key questions:

-   Which seasons have the highest probability of success?

-   Has the reason for ending an expedition changed in recent years?

-   Are there differences between expeditions led from Nepal, China or India?

-   Are there countries that are more present and more successful?

-   How much does the use of supplementary oxygen influence success?

## Import data

```{r}
################################################################################
### Cleaning data from the Himalayan Database ##################################
################################################################################


library(tidyverse)
library(foreign)
library(httr)
library(zip)
library(ggcorrplot)
library(formattable)

Sys.setlocale("LC_ALL", "C")  # resolve encoding problem

# Step 1: Setup URL 
url <- "https://www.himalayandatabase.com/downloads/Himalayan%20Database.zip"
zip_file <- tempfile(fileext = ".zip")
extracted_dir <- tempfile()

# Step 2: download ZIP and exctract
GET(url, write_disk(zip_file, overwrite = TRUE))
unzip(zip_file, exdir = extracted_dir)

dbf_file_peaks <- file.path(extracted_dir, "Himalayan Database", "HIMDATA", "peaks.DBF")
dbf_file_exped <- file.path(extracted_dir, "Himalayan Database", "HIMDATA", "exped.DBF")

# Step 5: read DBF and ignore warnings (already verified it)
peaks_data <- suppressWarnings(read.dbf(dbf_file_peaks, as.is = TRUE))
exped_data <- suppressWarnings(read.dbf(dbf_file_exped, as.is = TRUE))

# Step 6: convert into Tibble
peaks_temp <- as_tibble(peaks_data)
exped_temp <- as_tibble(exped_data)

unlink(c(zip_file, extracted_dir), recursive = TRUE)
```

```{r}
###_____________________________________________________________________________
### Add columns that recode some variables that are integers into character
### strings that describe their subject in the peaks dataset
###_____________________________________________________________________________

peaks_tidy <- peaks_temp |> 
  dplyr::mutate(HIMAL_FACTOR = case_when(
    HIMAL == 0  ~ "Unclassified",
    HIMAL == 1  ~ "Annapurna",
    HIMAL == 2  ~ "Api/Byas Risi/Guras",
    HIMAL == 3  ~ "Damodar",
    HIMAL == 4  ~ "Dhaulagiri",
    HIMAL == 5  ~ "Ganesh/Shringi",
    HIMAL == 6  ~ "Janak/Ohmi Kangri",
    HIMAL == 7  ~ "Jongsang",
    HIMAL == 8  ~ "Jugal",
    HIMAL == 9  ~ "Kangchenjunga/Simhalila",
    HIMAL == 10 ~ "Kanjiroba",
    HIMAL == 11 ~ "Kanti/Palchung",
    HIMAL == 12 ~ "Khumbu",
    HIMAL == 13 ~ "Langtang",
    HIMAL == 14 ~ "Makalu",
    HIMAL == 15 ~ "Manaslu/Mansiri",
    HIMAL == 16 ~ "Mukut/Mustang",
    HIMAL == 17 ~ "Nalakankar/Chandi/Changla",
    HIMAL == 18 ~ "Peri",
    HIMAL == 19 ~ "Rolwaling",
    HIMAL == 20 ~ "Saipal",
    TRUE ~ "Unknown" # Default case if a value doesn't match
  ),
  .after = HIMAL
  ) |> 
  mutate(REGION_FACTOR = case_when(
    REGION == 0 ~ "Unclassified",
    REGION == 1 ~ "Kangchenjunga-Janak",
    REGION == 2 ~ "Khumbu-Rolwaling-Makalu",
    REGION == 3 ~ "Langtang-Jugal",
    REGION == 4 ~ "Manaslu-Ganesh",
    REGION == 5 ~ "Annapurna-Damodar-Peri",
    REGION == 6 ~ "Dhaulagiri-Mukut",
    REGION == 7 ~ "Kanjiroba-Far West",
    TRUE        ~ "Unknown" # Default case if a value doesn't match
  ),
  .after = REGION
  ) |> 
  mutate(PHOST_FACTOR = case_when(
    PHOST == 0 ~ "Unclassified",
    PHOST == 1 ~ "Nepal only",
    PHOST == 2 ~ "China only",
    PHOST == 3 ~ "India only",
    PHOST == 4 ~ "Nepal & China",
    PHOST == 5 ~ "Nepal & India",
    PHOST == 6 ~ "Nepal, China & India",
    TRUE       ~ "Unknown" # Default case if a value doesn't match
  ),
  .after = PHOST
  ) |> 
  mutate(PSTATUS_FACTOR = case_when(
    PSTATUS == 0 ~ "Unknown",
    PSTATUS == 1 ~ "Unclimbed",
    PSTATUS == 2 ~ "Climbed",
    TRUE         ~ "Invalid" # Default case if a value doesn't match
  ),
  .after = PSTATUS
  ) |> 
  mutate(across(c(HIMAL_FACTOR, PHOST_FACTOR, PSTATUS_FACTOR), ~ factor(.)))

```

```{r}
exped_tidy <- exped_temp |> 
  dplyr::mutate(YEAR = as.numeric(trimws(YEAR))) |> 
  dplyr::filter(YEAR >= 2000) |> 
  dplyr::mutate(SEASON_FACTOR = case_when(
    SEASON == 0 ~ "Unknown",
    SEASON == 1 ~ "Spring",
    SEASON == 2 ~ "Summer",
    SEASON == 3 ~ "Autumn",
    SEASON == 4 ~ "Winter",
    TRUE ~ NA_character_
  ), .after = SEASON) |> 
  dplyr::mutate(HOST_FACTOR = case_when(
    HOST == 0 ~ "Unknown",
    HOST == 1 ~ "Nepal",
    HOST == 2 ~ "China",
    HOST == 3 ~ "India",
    TRUE ~ NA_character_
  ), .after = HOST) |> 
  dplyr::mutate(TERMREASON_FACTOR = case_when(
    TERMREASON == 0 ~ "Unknown",
    TERMREASON == 1 ~ "Success (main peak)",
    TERMREASON == 2 ~ "Success (subpeak, foresummit)",
    TERMREASON == 3 ~ "Success (claimed)",
    TERMREASON == 4 ~ "Bad weather (storms, high winds)",
    TERMREASON == 5 ~ "Bad conditions (deep snow, avalanching, falling ice, or rock)",
    TERMREASON == 6 ~ "Accident (death or serious injury)",
    TERMREASON == 7 ~ "Illness, AMS, exhaustion, or frostbite",
    TERMREASON == 8 ~ "Lack (or loss) of supplies, support or equipment",
    TERMREASON == 9 ~ "Lack of time",
    TERMREASON == 10 ~ "Route technically too difficult, lack of experience, strength, or motivation",
    TERMREASON == 11 ~ "Did not reach base camp",
    TERMREASON == 12 ~ "Did not attempt climb",
    TERMREASON == 13 ~ "Attempt rumored",
    TERMREASON == 14 ~ "Other"
  ), .after = TERMREASON)

```

```{r}
# not clean dataset
dim(exped_temp)
# clean dataset
dim(exped_tidy)
```

```{r}
# clean dataset
dim(peaks_tidy)
```

```{r}
head(exped_tidy)
```

```{r}
head(peaks_tidy)
```

```{r}
ggplot(exped_temp, aes(x = as.numeric(YEAR))) +
  geom_bar() +
  scale_x_continuous(
    breaks = seq(
      from = min(as.numeric(exped_temp$YEAR), na.rm = TRUE),
      to = max(as.numeric(exped_temp$YEAR), na.rm = TRUE),
      by = 20
    )
  ) +
  labs(
    title = "Number of Expeditions per Year",
    x = "Year",
    y = "Number of Expeditions"
  ) +
  theme_minimal()

```

## Data Cleaning

PEAK_TIDY DATASET DESCRIPTION

| Variable | Class | Description |
|-------------|-------------|----------------------------------------------|
| PEAKID | integer | Peak ID, the key field (unique identifier) for each record. |
| PKNAME | integer | Peak name. |
| PKNAME2 | integer | Peak name 2. |
| LOCATION | integer | Location of the peak. |
| HEIGHTM | integer | Peak height in meters. |
| HEIGHTF | integer | Peak height in feet. |
| HIMAL | integer | Mountain range identifier. See HIMAL_FACTOR for names. |
| HIMAL_FACTOR | integer | Mountain range name. |
| REGION | integer | Region identifier. |
| REGION_FACTOR | character | Region name. |
| OPEN | logical | Indicates whether the peak is open for expeditions. |
| UNLISTED | logical | Indicates whether the peak is unlisted in official records. |
| TREKKING | logical | Indicates whether the peak is a trekking peak. |
| TREKYEAR | integer | Year the peak was designated as a trekking peak. |
| RESTRICT | logical | Indicates whether the peak is restricted for climbing or access. |
| PHOST | integer | Primary host country identifier for the peak. See PHOST_FACTOR for names. |
| PHOST_FACTOR | character | Primary host country name corresponding to the PHOST identifier. |
| PSTATUS | integer | Climbing status identifier for the peak. See PSTATUS_FACTOR for names. |
| PSTATUS_FACTOR | character | Climbing status description corresponding to the PSTATUS identifier. |
| PEAKMEMO | character | Additional notes or remarks about the peak. |
| PYEAR | integer | Year of the first recorded climbing attempt on the peak. |
| PSEASON | character | Climbing season associated with the first recorded ascent or expedition. |
| PEXPID | integer | Expedition ID associated with the first recorded climb. |
| PSMTDATE | character | Date of the first successful summit. |
| PCOUNTRY | character | Country of origin for the expedition or climbers. |
| PSUMMITERS | integer | Number of climbers who first reached the summit. |
| PSMTNOTE | character | Notes or remarks related to the first successful summit. |
| REFERMEMO | character | Peak chronology references. |
| PHOTOMEMO | character | Peak photo references. |

EXPED_TIDY DATASET DESCRIPTION

| Variable | Class | Description |
|--------------|--------------|---------------------------------------------|
| EXPID | factor | Expedition ID. |
| PEAKID | factor | Peak ID. |
| YEAR | factor | Year of the expedition. |
| SEASON | integer | Season of the expedition as an integer. |
| SEASON_FACTOR | character | Season of the expedition converted to a descriptive name. |
| HOST | integer | Host country of the expedition as an integer. |
| HOST_FACTOR | character | Host country of the expedition converted to a descriptive name. |
| ROUTE1 | factor | Climbing route 1. |
| ROUTE2 | factor | Climbing route 2. |
| ROUTE3 | factor | Climbing route 3. |
| ROUTE4 | factor | Climbing route 4. |
| NATION | factor | Principal nationality. |
| LEADERS | factor | Leadership of the expedition. |
| SPONSOR | factor | Expedition sponsor / name. |
| SUCCESS1 | logical | Success on route 1? |
| SUCCESS2 | logical | Success on route 2? |
| SUCCESS3 | logical | Success on route 3? |
| SUCCESS4 | logical | Success on route 4? |
| ASCENT1 | factor | Ascent numbers for route 1. |
| ASCENT2 | factor | Ascent numbers for route 2. |
| ASCENT3 | factor | Ascent numbers for route 3. |
| ASCENT4 | factor | Ascent numbers for route 4. |
| CLAIMED | logical | Success claimed? |
| DISPUTED | logical | Success disputed? |
| COUNTRIES | factor | Other countries. |
| APPROACH | factor | Approach march. |
| BCDATE | date | Date arrived at base camp. |
| SMTDATE | date | Date reached summit. |
| SMTTIME | factor | Time reached summit. |
| SMTDAYS | integer | Number of days to summit / high-point. |
| TOTDAYS | integer | Total number of days. |
| TERMDATE | date | Date the expedition was terminated. |
| TERMREASON | integer | Reason the expedition was terminated as an integer. |
| TERMREASON_FACTOR | character | Reason the expedition was terminated converted to a descriptive name. |
| TERMNOTE | factor | Termination details. |
| HIGHPOINT | integer | Expedition high-point. |
| TRAVERSE | logical | Did the expedition traverse. |
| SKI | logical | Ski / snowboard descent? |
| PARAPENTE | logical | Parapente descent? |
| CAMPS | integer | Number of high camps above base camp. |
| ROPE | integer | Amount of fixed rope. |
| TOTMEMBERS | integer | Number of members. |
| SMTMEMBERS | integer | Number of members on summit. |
| MDEATHS | integer | Number of member deaths. |
| TOTHIRED | integer | Number of hired personnel (above base camp). |
| SMTHIRED | integer | Number of hired personnel on summit. |
| HDEATHS | integer | Number of hired personnel deaths. |
| NOHIRED | logical | No hired personnel used above base camp. |
| O2USED | logical | Oxygen used? |
| O2NONE | logical | Oxygen not used? |
| O2CLIMB | logical | Oxygen climbing? |
| O2DESCENT | logical | Oxygen descending? |
| O2SLEEP | logical | Oxygen sleeping? |
| O2MEDICAL | logical | Oxygen used medically? |
| O2TAKEN | logical | Oxygen taken, not used? |
| O2UNKWN | logical | Oxygen use unknown? |
| OTHERSMTS | factor | Other summits. |
| CAMPSITES | factor | Campsite details. |
| ROUTEMEMO | factor | Route details. |
| ACCIDENTS | factor | Accidents on the expedition. |
| ACHIEVMENT | factor | Achievements. |
| AGENCY | factor | Trekking agency. |
| COMRTE | logical | Commercial route? |
| STDRTE | logical | 8000m standard route? |
| PRIMRTE | logical | Route info with primary expedition? |
| PRIMMEM | logical | Member info with primary expedition? |
| PRIMREF | logical | Literature info with primary expedition? |
| PRIMID | factor | Primary expedition ID (if any). |
| CHKSUM | integer | Internal consistency check. |

### Handling NaN

At he moment the data contain a lot of NaN, I want to check If exist some columns that are principaly composed of missing values, if exists, we will remove them

```{r}
# 1. Compute % missing per column
miss_pct <- exped_tidy %>% 
  summarise_all(~ mean(is.na(.)) * 100) %>% 
  pivot_longer(everything(), names_to = "variable", values_to = "pct_missing")

# 2. Filter for columns with > 0% missing values
miss_pct_filtered <- miss_pct %>% 
  filter(pct_missing > 0)

# 3. Plot
miss_pct_filtered %>% 
  ggplot(aes(x = pct_missing, y = fct_reorder(variable, pct_missing))) +
  geom_col() +
  labs(
    x = "% missing",
    y = NULL,
    title = "Missing values per variable in exped_tidy (only > 0%)"
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, .05)))

```

```{r}
missing_pct <- round(colSums(is.na(exped_tidy)) / nrow(exped_tidy) * 100, 2)
missing_df <- data.frame(Column = names(missing_pct), MissingPercentage = missing_pct)

# Create a color-scaled table
formattable(
  missing_df[order(-missing_df$MissingPercentage), ],
  list(
    MissingPercentage = color_tile("white", "red")
  )
)
```

Now we are doing the same for the dataset peaks_tidy

```{r}
# 1. Compute % missing per column
miss_pct <- peaks_tidy %>% 
  summarise_all(~ mean(is.na(.)) * 100) %>% 
  pivot_longer(everything(), names_to = "variable", values_to = "pct_missing")

# 2. Filter for columns with > 0% missing values
miss_pct_filtered <- miss_pct %>% 
  filter(pct_missing > 0)

# 3. Plot
miss_pct_filtered %>% 
  ggplot(aes(x = pct_missing, y = fct_reorder(variable, pct_missing))) +
  geom_col() +
  labs(
    x = "% missing",
    y = NULL,
    title = "Missing values per variable in peaks_tidy (only > 0%)"
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, .05)))
```

```{r}
missing_pct <- round(colSums(is.na(peaks_tidy)) / nrow(peaks_tidy) * 100, 2)
missing_df <- data.frame(Column = names(missing_pct), MissingPercentage = missing_pct)

# Create a color-scaled table
formattable(
  missing_df[order(-missing_df$MissingPercentage), ],
  list(
    MissingPercentage = color_tile("white", "red")
  )
)
```

#### 🔴 Very High NA% (over 40%)

These columns have extremely high missing rates and are usually not useful unless they're absolutely critical:

-   `ASCENT3` 100.00%

-   `ASCENT4` 100.00%

-   `ROUTE4` 99.99%

-   `ROUTE3` 99.87%

-   `ASCENT2` 99.77%

-   `ROUTE2` 98.17%

-   `PRIMID` 95.32%

-   `ACHIEVMENT` 93.69%

-   `ASCENT1` 86.54%

-   `OTHERSMTS` 83.09%

-   `TERMNOTE` 62.30%

-   `COUNTRIES` 61.97%

-   `SMTTIME` 50.39%

-   `APPROACH` 45.40%

-   `REFERMEMO` 96.04%

-   `TREKYEAR` 93.96%

-   `PSMTNOTE` 84.17%

-   `PHOTOMEMO` 83.75%

-   `PKNAME2` 46.46%

-   `RESTRICT` 42.71%

We decided to drop all this columns because of the first analysis of the NaN values

```{r}
cols_to_drop_exped <- c(
  "ASCENT3", "ASCENT4", "ROUTE4", "ROUTE3", "ASCENT2", "ROUTE2", "PRIMID", "ACHIEVMENT",
  "ASCENT1", "OTHERSMTS", "TERMNOTE", "COUNTRIES", "SMTTIME", "APPROACH"
)

exped_tidy <- exped_tidy |> dplyr::select(-all_of(cols_to_drop_exped))

cols_to_drop_peaks <- c(
  "REFERMEMO", "TREKYEAR", "PSMTNOTE", "PHOTOMEMO", "PKNAME2", "RESTRICT"
)

peaks_tidy <- peaks_tidy |> dplyr::select(-all_of(cols_to_drop_peaks))

```

Duplicates:

```{r}
duplicated_rows_exped <- exped_tidy[duplicated(exped_tidy), ]
duplicated_rows_peaks <- peaks_tidy[duplicated(peaks_tidy), ]
nrow(duplicated_rows_exped)
nrow(duplicated_rows_peaks)
```

Join the two datasets

```{r}
exped_full <- exped_tidy |> 
  left_join(peaks_tidy, by = "PEAKID")
```

## Data Exploration (EDA)

Information about the

Overview of the dataset and the features

```{r}
sapply(exped_full, class) |> table()
```

```{r}
sapply(exped_full, class) |> sort()

```

Analysis of the numerical features:

```{r}
numeric_vars <- names(exped_tidy)[sapply(exped_tidy, is.numeric)]
print(numeric_vars)

```

```{r}
for (var in numeric_vars) {
  print(
    ggplot(exped_tidy, aes(x = .data[[var]])) +
      geom_histogram(bins = 30, fill = "steelblue", color = "white") +
      labs(title = paste("Distribution of", var), x = var, y = "Frequency")
  )
}
```

```{r}
cor_matrix <- exped_full |> 
  dplyr::select(all_of(numeric_vars)) |> 
  cor(use = "complete.obs")

library(ggcorrplot)
ggcorrplot::ggcorrplot(cor_matrix, lab = TRUE, lab_size = 3)
```

```{r}
library(GGally)
exped_full_num <- exped_full |> select(TOTDAYS, SMTDAYS, TOTMEMBERS, SMTMEMBERS, HIGHPOINT)
GGally::ggpairs(exped_full_num)
```

```{r}
categorical_vars <- names(exped_full)[sapply(exped_full, function(x) is.character(x) || is.factor(x))]

print(categorical_vars)

```

```{r}
ggplot(exped_full, aes(x = SEASON_FACTOR)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Distribution of expeditions per season", x = "Season", y = "Count")

```

```{r}
# Distribution by host
exped_full |> count(HOST_FACTOR) |> ggplot(aes(x = HOST_FACTOR, y = n)) +
  geom_bar(stat = "identity") + labs(title = "Distribution by host", x = "Host", y = "expedition")
```

```{r}
# Success per season
exped_tidy |> 
 mutate(success = ifelse(SUCCESS1, "Success", "Failure")) |> 
  count(SEASON_FACTOR, success) |> 
 ggplot(aes(x = SEASON_FACTOR, y = n, fill = success)) +
 geom_bar(stat = "identity", position = "dodge") +
 labs(title = "Successes and Failures by Season", x = "Season", y = "Number", fill = "Outcome")
```

```{r}
# Oxygen and success
exped_tidy |> 
 mutate(O2_USED = ifelse(O2USED, "O2 Used", "No O2")) |> 
 mutate(success = ifelse(SUCCESS1, "Success", "Failure")) |> 
  count(O2_USED, success) |> 
  ggplot(aes(x = O2_USED, y = n, fill = success)) +
 geom_bar(stat = "identity", position = "dodge") +
 labs(title = "Success in relation to Oxygen", x = "Oxygen", y = "Number", fill = "Result")

```

```{r}
# Boxplot giorni
ggplot(exped_tidy, aes(y = TOTDAYS)) + 
  geom_boxplot() + 
  labs(title = "Outliers in the total expedition days", y = "tot days")

```

```{r}

# outliers in members
exped_tidy |> filter(TOTMEMBERS > 30) |> select(EXPID, TOTMEMBERS)
```

```         
```

## Modeling

## Conclusion
