---
title: "R Notebook"
output:
  html_notebook
---

# Final project Data Projects & Hackathon 2

# Exploring Mountaineering Success in the Himalayas: An Analysis of Recent Expeditions (2000–2025)

Dataset: <https://github.com/rfordatascience/tidytuesday/tree/main/data/2025/2025-01-21>

## Introduction and problem setting

This project is based on the “Himalayan Database”, a detailed collection of all documented expeditions to the Nepalese Himalayas since 1905. In particular, we will analyse data on expeditions from 2010 onwards, in order to understand the factors influencing the success or failure of recent mountaineering missions.

The main objective is to explore what factors influence the outcome of an expedition. Some key questions:

-   Does the use of supplementary oxygen reduce the duration of expeditions?

-   Which features help expeditions reach the summit?

-   Are autumn expeditions more likely to fail than spring expeditions?

-   Does the use of supplementary oxygen increase the probability of success?

-   Are expeditions hosted by Nepal or China more likely to succeed?

These questions are particularly relevant given the rich structure of the dataset.

Understanding the role of **oxygen usage** can help assess both its practical value and potential risks in high-altitude climbing. Analyzing **altitude** and **duration** relationships provides insight into how physical challenges scale with mountain difficulty. Seasonal variation is a classic factor in expedition planning, and quantifying its impact on success rates can inform safer scheduling. Finally, exploring **leadership origin** may reflect differences in strategy, experience, or logistical capacity across countries, revealing potential cultural or regional patterns in expedition outcomes.

These questions are not only statistically approachable through regression models, but also contextually meaningful for both mountaineering professionals and enthusiasts.

## Import data

```{r}
library(tidyverse)
library(foreign)
library(httr)
library(zip)
library(ggcorrplot)
library(formattable)
library(VIM)
library(tidymodels)
library(glmnet)
library(infer)
```

```{r}
################################################################################
### Cleaning data from the Himalayan Database ##################################
################################################################################


Sys.setlocale("LC_ALL", "C")  # resolve encoding problem

# Step 1: Setup URL 
url <- "https://www.himalayandatabase.com/downloads/Himalayan%20Database.zip"
zip_file <- tempfile(fileext = ".zip")
extracted_dir <- tempfile()

# Step 2: download ZIP and exctract
GET(url, write_disk(zip_file, overwrite = TRUE))
unzip(zip_file, exdir = extracted_dir)

dbf_file_peaks <- file.path(extracted_dir, "Himalayan Database", "HIMDATA", "peaks.DBF")
dbf_file_exped <- file.path(extracted_dir, "Himalayan Database", "HIMDATA", "exped.DBF")

# Step 5: read DBF and ignore warnings (already verified it)
peaks_data <- suppressWarnings(read.dbf(dbf_file_peaks, as.is = TRUE))
exped_data <- suppressWarnings(read.dbf(dbf_file_exped, as.is = TRUE))

# Step 6: convert into Tibble
peaks_temp <- as_tibble(peaks_data)
exped_temp <- as_tibble(exped_data)

unlink(c(zip_file, extracted_dir), recursive = TRUE)
```

```{r}
###_____________________________________________________________________________
### Add columns that recode some variables that are integers into character
### strings that describe their subject in the peaks dataset
###_____________________________________________________________________________

peaks_tidy <- peaks_temp |> 
  dplyr::mutate(HIMAL_FACTOR = case_when(
    HIMAL == 0  ~ "Unclassified",
    HIMAL == 1  ~ "Annapurna",
    HIMAL == 2  ~ "Api/Byas Risi/Guras",
    HIMAL == 3  ~ "Damodar",
    HIMAL == 4  ~ "Dhaulagiri",
    HIMAL == 5  ~ "Ganesh/Shringi",
    HIMAL == 6  ~ "Janak/Ohmi Kangri",
    HIMAL == 7  ~ "Jongsang",
    HIMAL == 8  ~ "Jugal",
    HIMAL == 9  ~ "Kangchenjunga/Simhalila",
    HIMAL == 10 ~ "Kanjiroba",
    HIMAL == 11 ~ "Kanti/Palchung",
    HIMAL == 12 ~ "Khumbu",
    HIMAL == 13 ~ "Langtang",
    HIMAL == 14 ~ "Makalu",
    HIMAL == 15 ~ "Manaslu/Mansiri",
    HIMAL == 16 ~ "Mukut/Mustang",
    HIMAL == 17 ~ "Nalakankar/Chandi/Changla",
    HIMAL == 18 ~ "Peri",
    HIMAL == 19 ~ "Rolwaling",
    HIMAL == 20 ~ "Saipal",
    TRUE ~ "Unknown" # Default case if a value doesn't match
  ),
  .after = HIMAL
  ) |> 
  mutate(REGION_FACTOR = case_when(
    REGION == 0 ~ "Unclassified",
    REGION == 1 ~ "Kangchenjunga-Janak",
    REGION == 2 ~ "Khumbu-Rolwaling-Makalu",
    REGION == 3 ~ "Langtang-Jugal",
    REGION == 4 ~ "Manaslu-Ganesh",
    REGION == 5 ~ "Annapurna-Damodar-Peri",
    REGION == 6 ~ "Dhaulagiri-Mukut",
    REGION == 7 ~ "Kanjiroba-Far West",
    TRUE        ~ "Unknown" # Default case if a value doesn't match
  ),
  .after = REGION
  ) |> 
  mutate(PHOST_FACTOR = case_when(
    PHOST == 0 ~ "Unclassified",
    PHOST == 1 ~ "Nepal only",
    PHOST == 2 ~ "China only",
    PHOST == 3 ~ "India only",
    PHOST == 4 ~ "Nepal & China",
    PHOST == 5 ~ "Nepal & India",
    PHOST == 6 ~ "Nepal, China & India",
    TRUE       ~ "Unknown" # Default case if a value doesn't match
  ),
  .after = PHOST
  ) |> 
  mutate(PSTATUS_FACTOR = case_when(
    PSTATUS == 0 ~ "Unknown",
    PSTATUS == 1 ~ "Unclimbed",
    PSTATUS == 2 ~ "Climbed",
    TRUE         ~ "Invalid" # Default case if a value doesn't match
  ),
  .after = PSTATUS
  ) |> 
  mutate(across(c(HIMAL_FACTOR, PHOST_FACTOR, PSTATUS_FACTOR), ~ factor(.)))

```

```{r}
exped_tidy <- exped_temp |> 
  dplyr::mutate(YEAR = as.numeric(trimws(YEAR))) |> 
  dplyr::filter(YEAR >= 2000) |> 
  dplyr::mutate(SEASON_FACTOR = case_when(
    SEASON == 0 ~ "Unknown",
    SEASON == 1 ~ "Spring",
    SEASON == 2 ~ "Summer",
    SEASON == 3 ~ "Autumn",
    SEASON == 4 ~ "Winter",
    TRUE ~ NA_character_
  ), .after = SEASON) |> 
  dplyr::mutate(HOST_FACTOR = case_when(
    HOST == 0 ~ "Unknown",
    HOST == 1 ~ "Nepal",
    HOST == 2 ~ "China",
    HOST == 3 ~ "India",
    TRUE ~ NA_character_
  ), .after = HOST) |> 
  dplyr::mutate(TERMREASON_FACTOR = case_when(
    TERMREASON == 0 ~ "Unknown",
    TERMREASON == 1 ~ "Success (main peak)",
    TERMREASON == 2 ~ "Success (subpeak, foresummit)",
    TERMREASON == 3 ~ "Success (claimed)",
    TERMREASON == 4 ~ "Bad weather (storms, high winds)",
    TERMREASON == 5 ~ "Bad conditions (deep snow, avalanching, falling ice, or rock)",
    TERMREASON == 6 ~ "Accident (death or serious injury)",
    TERMREASON == 7 ~ "Illness, AMS, exhaustion, or frostbite",
    TERMREASON == 8 ~ "Lack (or loss) of supplies, support or equipment",
    TERMREASON == 9 ~ "Lack of time",
    TERMREASON == 10 ~ "Route technically too difficult, lack of experience, strength, or motivation",
    TERMREASON == 11 ~ "Did not reach base camp",
    TERMREASON == 12 ~ "Did not attempt climb",
    TERMREASON == 13 ~ "Attempt rumored",
    TERMREASON == 14 ~ "Other"
  ), .after = TERMREASON)

```

```{r}
# not clean dataset
dim(exped_temp)
# clean dataset
dim(exped_tidy)
```

```{r}
# clean dataset
dim(peaks_tidy)
```

```{r}
head(exped_tidy)
```

```{r}
head(peaks_tidy)
```

exploring the number of shipments by year, we realised that shipments started to increase exponentially from around the year 2000. therefore we decided to look at the data from this date onwards

```{r}
ggplot(exped_temp, aes(x = as.numeric(YEAR))) +
  geom_bar() +
  scale_x_continuous(
    breaks = seq(
      from = min(as.numeric(exped_temp$YEAR), na.rm = TRUE),
      to = max(as.numeric(exped_temp$YEAR), na.rm = TRUE),
      by = 20
    )
  ) +
  labs(
    title = "Number of Expeditions per Year",
    x = "Year",
    y = "Number of Expeditions"
  ) +
  theme_minimal()

```

## Data Cleaning and Preprocessing

The dataset we will use at the end of the data cleaning and preprocessing is a **merged version** of the peaks_tidy and exped_tidy datasets, filtered to include only expeditions from the year 2000 onwards. The main steps of cleaning included:

-   A detailed **description of available features**, distinguishing between numeric, categorical, and logical variables.

-   Handling of missing values:

    -   **Features with more than 40% missing values were discarded**.

    -   For the remaining ones, missing values are being addressed through appropriate **data imputation techniques** (to be finalized, e.g., median/mode for numerical/categorical, or domain-specific strategies).

    Planned additions:

    -   Identify and address duplicate entries.

    -   Standardize categorical levels (e.g., for country names, reasons for termination, etc.).

PEAK_TIDY DATASET DESCRIPTION

| Variable | Class | Description |
|------------------------|------------------------|------------------------|
| PEAKID | integer | Peak ID, the key field (unique identifier) for each record. |
| PKNAME | integer | Peak name. |
| PKNAME2 | integer | Peak name 2. |
| LOCATION | integer | Location of the peak. |
| HEIGHTM | integer | Peak height in meters. |
| HEIGHTF | integer | Peak height in feet. |
| HIMAL | integer | Mountain range identifier. See HIMAL_FACTOR for names. |
| HIMAL_FACTOR | integer | Mountain range name. |
| REGION | integer | Region identifier. |
| REGION_FACTOR | character | Region name. |
| OPEN | logical | Indicates whether the peak is open for expeditions. |
| UNLISTED | logical | Indicates whether the peak is unlisted in official records. |
| TREKKING | logical | Indicates whether the peak is a trekking peak. |
| TREKYEAR | integer | Year the peak was designated as a trekking peak. |
| RESTRICT | logical | Indicates whether the peak is restricted for climbing or access. |
| PHOST | integer | Primary host country identifier for the peak. See PHOST_FACTOR for names. |
| PHOST_FACTOR | character | Primary host country name corresponding to the PHOST identifier. |
| PSTATUS | integer | Climbing status identifier for the peak. See PSTATUS_FACTOR for names. |
| PSTATUS_FACTOR | character | Climbing status description corresponding to the PSTATUS identifier. |
| PEAKMEMO | character | Additional notes or remarks about the peak. |
| PYEAR | integer | Year of the first recorded climbing attempt on the peak. |
| PSEASON | character | Climbing season associated with the first recorded ascent or expedition. |
| PEXPID | integer | Expedition ID associated with the first recorded climb. |
| PSMTDATE | character | Date of the first successful summit. |
| PCOUNTRY | character | Country of origin for the expedition or climbers. |
| PSUMMITERS | integer | Number of climbers who first reached the summit. |
| PSMTNOTE | character | Notes or remarks related to the first successful summit. |
| REFERMEMO | character | Peak chronology references. |
| PHOTOMEMO | character | Peak photo references. |

EXPED_TIDY DATASET DESCRIPTION

| Variable | Class | Description |
|------------------------|------------------------|------------------------|
| EXPID | factor | Expedition ID. |
| PEAKID | factor | Peak ID. |
| YEAR | factor | Year of the expedition. |
| SEASON | integer | Season of the expedition as an integer. |
| SEASON_FACTOR | character | Season of the expedition converted to a descriptive name. |
| HOST | integer | Host country of the expedition as an integer. |
| HOST_FACTOR | character | Host country of the expedition converted to a descriptive name. |
| ROUTE1 | factor | Climbing route 1. |
| ROUTE2 | factor | Climbing route 2. |
| ROUTE3 | factor | Climbing route 3. |
| ROUTE4 | factor | Climbing route 4. |
| NATION | factor | Principal nationality. |
| LEADERS | factor | Leadership of the expedition. |
| SPONSOR | factor | Expedition sponsor / name. |
| SUCCESS1 | logical | Success on route 1? |
| SUCCESS2 | logical | Success on route 2? |
| SUCCESS3 | logical | Success on route 3? |
| SUCCESS4 | logical | Success on route 4? |
| ASCENT1 | factor | Ascent numbers for route 1. |
| ASCENT2 | factor | Ascent numbers for route 2. |
| ASCENT3 | factor | Ascent numbers for route 3. |
| ASCENT4 | factor | Ascent numbers for route 4. |
| CLAIMED | logical | Success claimed? |
| DISPUTED | logical | Success disputed? |
| COUNTRIES | factor | Other countries. |
| APPROACH | factor | Approach march. |
| BCDATE | date | Date arrived at base camp. |
| SMTDATE | date | Date reached summit. |
| SMTTIME | factor | Time reached summit. |
| SMTDAYS | integer | Number of days to summit / high-point. |
| TOTDAYS | integer | Total number of days. |
| TERMDATE | date | Date the expedition was terminated. |
| TERMREASON | integer | Reason the expedition was terminated as an integer. |
| TERMREASON_FACTOR | character | Reason the expedition was terminated converted to a descriptive name. |
| TERMNOTE | factor | Termination details. |
| HIGHPOINT | integer | Expedition high-point. |
| TRAVERSE | logical | Did the expedition traverse. |
| SKI | logical | Ski / snowboard descent? |
| PARAPENTE | logical | Parapente descent? |
| CAMPS | integer | Number of high camps above base camp. |
| ROPE | integer | Amount of fixed rope. |
| TOTMEMBERS | integer | Number of members. |
| SMTMEMBERS | integer | Number of members on summit. |
| MDEATHS | integer | Number of member deaths. |
| TOTHIRED | integer | Number of hired personnel (above base camp). |
| SMTHIRED | integer | Number of hired personnel on summit. |
| HDEATHS | integer | Number of hired personnel deaths. |
| NOHIRED | logical | No hired personnel used above base camp. |
| O2USED | logical | Oxygen used? |
| O2NONE | logical | Oxygen not used? |
| O2CLIMB | logical | Oxygen climbing? |
| O2DESCENT | logical | Oxygen descending? |
| O2SLEEP | logical | Oxygen sleeping? |
| O2MEDICAL | logical | Oxygen used medically? |
| O2TAKEN | logical | Oxygen taken, not used? |
| O2UNKWN | logical | Oxygen use unknown? |
| OTHERSMTS | factor | Other summits. |
| CAMPSITES | factor | Campsite details. |
| ROUTEMEMO | factor | Route details. |
| ACCIDENTS | factor | Accidents on the expedition. |
| ACHIEVMENT | factor | Achievements. |
| AGENCY | factor | Trekking agency. |
| COMRTE | logical | Commercial route? |
| STDRTE | logical | 8000m standard route? |
| PRIMRTE | logical | Route info with primary expedition? |
| PRIMMEM | logical | Member info with primary expedition? |
| PRIMREF | logical | Literature info with primary expedition? |
| PRIMID | factor | Primary expedition ID (if any). |
| CHKSUM | integer | Internal consistency check. |

### Handling NaN

At he moment the data contain a lot of NaN, I want to check If exist some columns that are principaly composed of missing values, if exists, we will remove them

```{r}
# 1. Compute % missing per column
miss_pct <- exped_tidy %>% 
  summarise_all(~ mean(is.na(.)) * 100) %>% 
  pivot_longer(everything(), names_to = "variable", values_to = "pct_missing")

# 2. Filter for columns with > 0% missing values
miss_pct_filtered <- miss_pct %>% 
  filter(pct_missing > 0)

# 3. Plot
miss_pct_filtered %>% 
  ggplot(aes(x = pct_missing, y = fct_reorder(variable, pct_missing))) +
  geom_col() +
  labs(
    x = "% missing",
    y = NULL,
    title = "Missing values per variable in exped_tidy (only > 0%)"
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, .05)))

```

```{r}
missing_pct <- round(colSums(is.na(exped_tidy)) / nrow(exped_tidy) * 100, 2)
missing_df <- data.frame(Column = names(missing_pct), MissingPercentage = missing_pct)

# Create a color-scaled table
formattable(
  missing_df[order(-missing_df$MissingPercentage), ],
  list(
    MissingPercentage = color_tile("white", "red")
  )
)
```

Now we are doing the same for the dataset peaks_tidy

```{r}
# 1. Compute % missing per column
miss_pct <- peaks_tidy %>% 
  summarise_all(~ mean(is.na(.)) * 100) %>% 
  pivot_longer(everything(), names_to = "variable", values_to = "pct_missing")

# 2. Filter for columns with > 0% missing values
miss_pct_filtered <- miss_pct %>% 
  filter(pct_missing > 0)

# 3. Plot
miss_pct_filtered %>% 
  ggplot(aes(x = pct_missing, y = fct_reorder(variable, pct_missing))) +
  geom_col() +
  labs(
    x = "% missing",
    y = NULL,
    title = "Missing values per variable in peaks_tidy (only > 0%)"
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, .05)))
```

```{r}
missing_pct <- round(colSums(is.na(peaks_tidy)) / nrow(peaks_tidy) * 100, 2)
missing_df <- data.frame(Column = names(missing_pct), MissingPercentage = missing_pct)

# Create a color-scaled table
formattable(
  missing_df[order(-missing_df$MissingPercentage), ],
  list(
    MissingPercentage = color_tile("white", "red")
  )
)
```

#### Very High NA% (over 40%)

These columns have extremely high missing rates and are usually not useful unless they're absolutely critical:

-   `ASCENT3` 100.00%

-   `ASCENT4` 100.00%

-   `ROUTE4` 99.99%

-   `ROUTE3` 99.87%

-   `ASCENT2` 99.77%

-   `ROUTE2` 98.17%

-   `PRIMID` 95.32%

-   `ACHIEVMENT` 93.69%

-   `ASCENT1` 86.54%

-   `OTHERSMTS` 83.09%

-   `TERMNOTE` 62.30%

-   `COUNTRIES` 61.97%

-   `SMTTIME` 50.39%

-   `APPROACH` 45.40%

-   `REFERMEMO` 96.04%

-   `TREKYEAR` 93.96%

-   `PSMTNOTE` 84.17%

-   `PHOTOMEMO` 83.75%

-   `PKNAME2` 46.46%

-   `RESTRICT` 42.71%

We decided to drop all this columns because of the first analysis of the NaN values

```{r}
cols_to_drop_exped <- c(
  "ASCENT3", "ASCENT4", "ROUTE4", "ROUTE3", "ASCENT2", "ROUTE2", "PRIMID", "ACHIEVMENT",
  "ASCENT1", "OTHERSMTS", "TERMNOTE", "COUNTRIES", "SMTTIME", "APPROACH"
)

exped_tidy <- exped_tidy |> dplyr::select(-all_of(cols_to_drop_exped))

cols_to_drop_peaks <- c(
  "REFERMEMO", "TREKYEAR", "PSMTNOTE", "PHOTOMEMO", "PKNAME2", "RESTRICT"
)

peaks_tidy <- peaks_tidy |> dplyr::select(-all_of(cols_to_drop_peaks))

```

### Handling features that aren't informative

After handling columns with excessive missing values, we proceed to remove additional variables that are not informative for modeling or analysis.

These columns fall into one or more of the following categories: -

-   Descriptive text fields (e.g., notes, comments, or memos) that are unique and not analytically useful

-   Redundant variables where a clearer human-readable factor version already exists

-   High-cardinality identifiers (e.g., names or sponsors) that are not relevant to our research questions

This step is crucial to reduce noise, avoid overfitting in machine learning models, and simplify our dataset for exploration and modeling.

```{r}
# Count unique values per column
unique_counts <- sapply(exped_tidy, function(x) length(unique(x)))

# Sort descending
sort(unique_counts, decreasing = TRUE)
```

```{r}
columns_to_drop_peaks <- c(
  "PKNAME2", "HEIGHTF", "HIMAL_FACTOR", "REGION_FACTOR", "OPEN",
  "UNLISTED", "TREKKING", "TREKYEAR", "RESTRICT", "PHOST_FACTOR",
  "PSTATUS_FACTOR", "PEAKMEMO", "PYEAR", "PSEASON", "PEXPID",
  "PSMTDATE", "PCOUNTRY", "PSUMMITERS", "PSMTNOTE", "REFERMEMO",
  "PHOTOMEMO"
)

columns_to_drop_exped <- c(
  "SPONSOR", "ASCENT1", "ASCENT2", "ASCENT3", "ASCENT4", "APPROACH", "TERMREASON_FACTOR", "TERMNOTE", "SKI", "PARAPENTE", "O2NONE","CAMPSITES", "ROUTEMEMO","ACHIEVMENT", "PRIMRTE", "PRIMMEM", "PRIMREF", "PRIMID", "CHKSUM", "ACCIDENTS", "AGENCY", "OTHERSMTS", "LEADERS", "NATION", "COUNTRIES"
)

peaks_tidy <- peaks_tidy |> select(-any_of(columns_to_drop_peaks))

exped_tidy <- exped_tidy |> select(-any_of(c(columns_to_drop_exped)))

```

### Missing values imputation

To ensure our dataset is complete and ready for modeling, we apply appropriate missing value imputation.

Different strategies are used based on variable types:

-   For numerical variables: we use median imputation to avoid distortion by outliers.

-   For categorical variables: we impute with `"Missing"` or the most frequent category (mode). -

-   For logical variables: missing values are assumed to be `FALSE` when safe, or flagged as `"Unknown"` where interpretation is unclear.

This step ensures that we do not lose rows due to NAs during model fitting.

```{r}
# Imputation - exped_tidy
numeric_vars_exped <- names(exped_tidy)[sapply(exped_tidy, is.numeric)]
exped_numeric <- exped_tidy[, numeric_vars_exped]

# KNN Imputation for numeric vars
exped_numeric_imputed <- kNN(exped_numeric, k = 5, imp_var = FALSE)
exped_tidy[numeric_vars_exped] <- exped_numeric_imputed

# Categorical
cat_vars_exped <- names(exped_tidy)[sapply(exped_tidy, function(x) is.character(x) || is.factor(x))]
for (col in cat_vars_exped) {
  exped_tidy[[col]] <- as.character(exped_tidy[[col]])
  exped_tidy[[col]][is.na(exped_tidy[[col]])] <- "Missing"
  exped_tidy[[col]] <- as.factor(exped_tidy[[col]])
}

# Logical
log_vars_exped <- names(exped_tidy)[sapply(exped_tidy, is.logical)]
for (col in log_vars_exped) {
  exped_tidy[[col]][is.na(exped_tidy[[col]])] <- FALSE
}
```

```{r}
# Imputation - peaks_tidy
numeric_vars_peaks <- names(peaks_tidy)[sapply(peaks_tidy, is.numeric)]
peaks_numeric <- peaks_tidy[, numeric_vars_peaks]

# KNN Imputation
peaks_numeric_imputed <- kNN(peaks_numeric, k = 5, imp_var = FALSE)
peaks_tidy[numeric_vars_peaks] <- peaks_numeric_imputed

# Categorical
cat_vars_peaks <- names(peaks_tidy)[sapply(peaks_tidy, function(x) is.character(x) || is.factor(x))]
for (col in cat_vars_peaks) {
  peaks_tidy[[col]] <- as.character(peaks_tidy[[col]])
  peaks_tidy[[col]][is.na(peaks_tidy[[col]])] <- "Missing"
  peaks_tidy[[col]] <- as.factor(peaks_tidy[[col]])
}

# Logical
log_vars_peaks <- names(peaks_tidy)[sapply(peaks_tidy, is.logical)]
for (col in log_vars_peaks) {
  peaks_tidy[[col]][is.na(peaks_tidy[[col]])] <- FALSE
}
```

```{r}
sum(is.na(exped_tidy)) 
sum(is.na(peaks_tidy))   
```

Every imputation where done correctly, but we have still some missing datas. So let's explore it again

```{r}
na_counts <- colSums(is.na(exped_tidy))
na_counts[na_counts > 0]
```

```{r}
library(lubridate)

exped_tidy <- exped_tidy |> mutate(
  DAYS_TO_END = as.numeric(TERMDATE - BCDATE),
  DAY_OF_YEAR_START = yday(BCDATE)
)
```

```{r}
exped_tidy <- exped_tidy |> select(-BCDATE, -SMTDATE, -TERMDATE)
```

```{r}
colSums(is.na(exped_tidy[c("DAYS_TO_END", "DAY_OF_YEAR_START")]))

# Imputation con KNN se necessario

to_impute <- exped_tidy[, c("DAYS_TO_END", "DAY_OF_YEAR_START")]
to_impute_knn <- kNN(to_impute, k = 5, imp_var = FALSE)
exped_tidy[, c("DAYS_TO_END", "DAY_OF_YEAR_START")] <- to_impute_knn
```

```{r}
# Now should be both 0
sum(is.na(exped_tidy)) 
sum(is.na(peaks_tidy))   
```

We transformed date fields into numeric variables to preserve meaningful temporal information while making them usable in modeling. Specifically, we computed:

-   `DAYS_TO_END`: days from base camp arrival to expedition termination

-   `DAY_OF_YEAR_START`: expedition start day of the year

We removed the original date columns and applied KNN imputation (k = 5) on the derived variables to handle remaining missing values.

### Duplicates

```{r}
duplicated_rows_exped <- exped_tidy[duplicated(exped_tidy), ]
duplicated_rows_peaks <- peaks_tidy[duplicated(peaks_tidy), ]
nrow(duplicated_rows_exped)
nrow(duplicated_rows_peaks)
```

### Join the two datasets

```{r}
exped_full <- exped_tidy |> 
  left_join(peaks_tidy, by = "PEAKID")
```

### Encoding

To make our dataset suitable for modeling, we encoded all categorical variables as factors. This allows R's linear and logistic regression functions (`lm()`, `glm()`) to automatically generate appropriate dummy variables internally.

```{r}
# Convert character columns to factor
exped_full <- exped_full |> mutate(
  across(where(is.character), as.factor)
)
```

```{r}
# List categorical variables
cat_vars <- names(exped_full)[sapply(exped_full, is.factor)]

```

### Scaling

We applied standard scaling to all continuous numeric variables, transforming them to have a mean of 0 and a standard deviation of 1. This step is important for models like linear and logistic regression, where variables on very different scales can distort the estimation of coefficients.

We excluded logical and categorical variables from this process.

```{r}
# Identify right-skewed predictors (excluding HIGHPOINT)
right_skewed_vars <- c(
  "SMTDAYS", "TOTDAYS", "CAMPS", "ROPE", 
  "TOTMEMBERS", "SMTMEMBERS", "MDEATHS", 
  "TOTHIRED", "SMTHIRED", "HDEATHS"
)

# Apply log1p transformation to selected predictors only
exped_full <- exped_full |>
  mutate(across(all_of(right_skewed_vars), log1p))
```

```{r}
# Trova le variabili numeriche
numeric_vars <- names(exped_full)[sapply(exped_full, is.numeric)]

# Conta i valori unici
unique_counts <- sapply(exped_full[numeric_vars], function(x) length(unique(x)))

# Mostra le variabili con <= 10 valori unici (sospette categorie)
sort(unique_counts[unique_counts <= 10])

```

```{r}
# Variables to scale (excluding binary and categorical-encoded ones)
numeric_vars <- names(exped_full)[sapply(exped_full, is.numeric)]

# Exclude vars we don't want to scale
to_exclude <- c("SUCCESS1", "O2USED", "O2CLIMB", "O2DESCENT", "O2SLEEP",
                "O2MEDICAL", "O2TAKEN", "O2UNKWN", "NOHIRED",
                "SEASON", "HOST", "TERMREASON", "REGION", "PHOST", "PSTATUS", "HIMAL", "HIGHPOINT", "HEIGHTM")

vars_to_scale <- setdiff(numeric_vars, to_exclude)

# Apply z-score standardization
exped_full[vars_to_scale] <- scale(exped_full[vars_to_scale])
```

## Data Exploration (EDA)

Currently underway:

-   Summary statistics and distribution plots for **numerical features**.

-   Planned analysis for:

    -   **Categorical features** (e.g., country, season, termination reason)

    -   **Logical features** (e.g., use of oxygen, success/failure).

-   Exploratory visualizations to support modeling choices (e.g., boxplots, bar charts, correlation heatmaps).

Information about the

Overview of the dataset and the features

```{r}
sapply(exped_full, class) |> table()
```

```{r}
sapply(exped_full, class) |> sort()

```

Analysis of the numerical features:

```{r}
numeric_vars <- names(exped_full)[sapply(exped_full, is.numeric)]
print(numeric_vars)

```

```{r}
for (var in numeric_vars) {
  print(
    ggplot(exped_full, aes(x = .data[[var]])) +
      geom_histogram(bins = 30, fill = "steelblue", color = "white") +
      labs(title = paste("Distribution of", var), x = var, y = "Frequency")
  )
}
```

```{r}
cor_matrix <- exped_full |> 
  dplyr::select(all_of(numeric_vars)) |> 
  cor(use = "complete.obs")

library(ggcorrplot)
ggcorrplot::ggcorrplot(cor_matrix, lab = TRUE, lab_size = 3)
```

```{r}
exped_full_num <- exped_full |> select(TOTDAYS, SMTDAYS, TOTMEMBERS, SMTMEMBERS, HIGHPOINT)
GGally::ggpairs(exped_full_num)
```

```{r}
categorical_vars <- names(exped_full)[sapply(exped_full, function(x) is.character(x) || is.factor(x))]

print(categorical_vars)

```

```{r}
ggplot(exped_full, aes(x = SEASON_FACTOR)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Distribution of expeditions per season", x = "Season", y = "Count")

```

```{r}
# Distribution by host
exped_full |> count(HOST_FACTOR) |> ggplot(aes(x = HOST_FACTOR, y = n)) +
  geom_bar(stat = "identity") + labs(title = "Distribution by host", x = "Host", y = "expedition")
```

```{r}
# Success per season
exped_full |> 
 mutate(success = ifelse(SUCCESS1, "Success", "Failure")) |> 
  count(SEASON_FACTOR, success) |> 
 ggplot(aes(x = SEASON_FACTOR, y = n, fill = success)) +
 geom_bar(stat = "identity", position = "dodge") +
 labs(title = "Successes and Failures by Season", x = "Season", y = "Number", fill = "Outcome")
```

```{r}
# Oxygen and success
exped_full |> 
 mutate(O2_USED = ifelse(O2USED, "O2 Used", "No O2")) |> 
 mutate(success = ifelse(SUCCESS1, "Success", "Failure")) |> 
  count(O2_USED, success) |> 
  ggplot(aes(x = O2_USED, y = n, fill = success)) +
 geom_bar(stat = "identity", position = "dodge") +
 labs(title = "Success in relation to Oxygen", x = "Oxygen", y = "Number", fill = "Result")

```

```{r}
# Boxplot giorni
ggplot(exped_full, aes(y = TOTDAYS)) + 
  geom_boxplot() + 
  labs(title = "Outliers in the total expedition days", y = "tot days")

```

## Modeling

After conducting a detailed exploratory data analysis on some interesting features of our datasets, we now move to the modeling phase. The goal of this section is to address key questions regarding expedition outcomes using statistical models introduced during the course.

The exploratory phase revealed several factors that may influence expedition success, such as the use of supplementary oxygen, the hosting country, the season, duration, and peak altitude. However, to rigorously assess the impact of each variable, we need to build models that can isolate and quantify these effects.

### Regression

#### 1. Does the use of supplementary oxygen reduce the duration of expeditions?

In this section, we aim to investigate whether the use of supplementary oxygen (`O2USED`) has a measurable effect on the total duration of Himalayan expeditions (`TOTDAYS`). This is a regression problem, as the response variable is continuous.

The variable `O2USED` is binary (`TRUE`/`FALSE`), indicating whether oxygen was used during the expedition. Our working hypothesis is that expeditions using oxygen may have shorter durations, as supplementary oxygen could improve endurance and reduce fatigue at high altitudes.

Understanding this relationship could offer valuable insights into the logistical and physiological implications of oxygen use in high-altitude mountaineering.

We begin with some exploratory visualizations and summary statistics, and then fit a linear regression model to test the association between oxygen use and expedition duration.

```{r}
exped_full %>%
  group_by(O2USED) %>%
  summarise(
    count = n(),
    mean_days = mean(TOTDAYS),
    sd_days = sd(TOTDAYS),
    median_days = median(TOTDAYS)
  )
```

We observe that:

-   Expeditions **without oxygen** (`O2USED = FALSE`) had a **mean duration** of approximately **-0.18** and a **median** of about **0.15**

-   Expeditions **with oxygen** (`O2USED = TRUE`) had a **mean duration** of about **0.34** and a **median** of **0.77**

Now let's see it in a graphically way:

```{r}
ggplot(exped_full, aes(x = O2USED, y = TOTDAYS, fill = O2USED)) +
  geom_boxplot() +
  labs(title = "Expedition Duration by Supplementary Oxygen Use",
       x = "Oxygen Used",
       y = "Total Days") +
  theme_minimal()

```

The boxplot confirms the trend observed in the summary statistics:\

**Expeditions using supplementary oxygen tend to last longer** than those that don’t.

To statistically assess whether supplementary oxygen affects expedition duration, we test the following hypotheses:

-   **Null hypothesis (H₀):** There is no difference in average expedition duration (`TOTDAYS`) between expeditions that used supplementary oxygen (`O2USED = TRUE`) and those that did not.
-   **Alternative hypothesis (H₁):** Expeditions that used oxygen had a different average duration than those that did not.Point estimates and summary statistics

We use a two-sample Welch's t-test (unequal variance assumed).

```{r}
t.test(TOTDAYS ~ O2USED, data = exped_full, var.equal = FALSE)
```

We performed a Welch two-sample t-test to compare the average expedition duration between groups using and not using supplementary oxygen.

-   The test yields a **highly significant result** (p-value \< 2.2e-16), with a **95% confidence interval** for the difference in means between **-0.56 and -0.47**.

-   This confirms that there is a **statistically significant difference** in average duration between the two groups.

Interestingly, expeditions **using oxygen actually last longer** on average, contradicting the idea that oxygen shortens climbs. This likely reflects that oxygen is more often used on longer, more demanding expeditions.

```{r}
# Bootstrap difference in means
set.seed(123)
oxygen_infer <- exped_full %>%
  specify(TOTDAYS ~ O2USED) %>%
  generate(reps = 1000, type = "bootstrap") %>%
  calculate(stat = "diff in means", order = c(TRUE, FALSE))

# Confidence interval
get_confidence_interval(oxygen_infer, level = 0.95, type = "percentile")

# Plot the bootstrap distribution
visualize(oxygen_infer) +
  geom_vline(xintercept = mean(oxygen_infer$stat), linetype = "dashed")


```

To confirm the result from the t-test, we applied a bootstrap approach to estimate the difference in mean expedition duration between the two groups (`O2USED = TRUE` vs `FALSE`).

The simulated bootstrap distribution is centered around **0.50**, reinforcing the finding that expeditions using supplementary oxygen tend to be longer.

This non-parametric method supports the conclusion that the observed difference is not due to random variation.

**Linear Regression**

We model the total expedition duration as:

$$
\text{TOTDAYS} = \beta_0 + \beta_1 \cdot \text{O2USED}
$$

Where:

-   $\beta_0$ is the **mean duration** of expeditions **without supplementary oxygen** (`O2USED = FALSE`)
-   $\beta_1$ is the **average difference in duration** between expeditions that **used oxygen** and those that did not

We model the total duration of expeditions (`TOTDAYS`) as a function of oxygen use (`O2USED`), using simple linear regression.

This model helps quantify the expected change in duration when oxygen is used, holding all else constant. Since oxygen use is binary, the coefficient represents the average difference in days between the two groups.

```{r}
# Fit linear regression
lm_oxy <- lm(TOTDAYS ~ O2USED, data = exped_full)

# Summary of model
summary(lm_oxy)
```

Residuals diagnostics

```{r}
# Residual plots
library(performance)
check_model(lm_oxy)
```

```{r}
exped_full %>%
  mutate(pred = predict(lm_oxy, newdata = exped_full)) %>%
  metrics(truth = TOTDAYS, estimate = pred)
```

The linear regression confirms that expeditions using supplementary oxygen are, on average, **0.51 standard units longer** than those that do not—holding all else constant. This effect is statistically significant (*p* \< 0.001).

However, the model’s R² is **very low (\~0.06)**, meaning oxygen use alone explains only a small portion of the variation in expedition duration. This suggests that many other factors likely influence duration, which this simple model does not capture, for this reason, a simple linear regression is not sufficient to capture the complexity behind the duration of a successful expedition.

The residual plots show some deviation from ideal assumptions, especially regarding homoscedasticity and normality. Despite this, the model is still useful as an initial approximation of the association.

#### 2. Which features help expeditions reach the summit?

In this section, we aim to understand which factors are most strongly associated with the performance of mountaineering expeditions, measured by the highest point reached (`HIGHPOINT`). This is a regression task with a continuous response variable.

Unlike the previous question, here we include multiple predictors to account for the complexity of an expedition. These include:

-   Team size (`TOTMEMBERS`, `SMTMEMBERS`)

-   Hired staff (`TOTHIRED`, `SMTHIRED`)

-   Use of supplementary oxygen (`O2USED`, `O2CLIMB`)

-   Climbing season (`SEASON`)

-   Hosting country (`HOST`)

-   Logistic factors such as camps, ropes, and support metrics (`CAMPS`, `ROPE`, `COMRTE`, `STDRTE`)

```{r}
# Filter only complete cases for modeling
model_data <- exped_full %>%
  select(HIGHPOINT, O2USED, TOTMEMBERS, SMTMEMBERS, TOTHIRED, SMTHIRED, O2CLIMB, 
         SEASON, HOST, CAMPS, ROPE, COMRTE, STDRTE, HOST_FACTOR) %>%
  drop_na()
```

```{r}
summary(model_data$HIGHPOINT)

ggplot(model_data, aes(x = HIGHPOINT)) +
  geom_histogram(bins = 30, fill = "steelblue") +
  labs(title = "Distribution of Highpoint") +
  theme_minimal()
```

### Exploring the Target Variable

We begin by exploring the distribution of the outcome variable `HIGHPOINT`, which represents the highest altitude reached during an expedition. The histogram shows a bimodal and skewed pattern:

-   A large number of expeditions reach very low altitudes (left side), possibly indicating early turnarounds or failures.

-   The right tail represents expeditions that approached or reached the summit, with noticeable peaks near 7000–8000 meters.

**Does oxygen use affect highpoint?**

To assess whether the use of supplementary oxygen has a significant effect on expedition performance, we tested for a difference in `HIGHPOINT` between groups that used oxygen and those that did not.

-   **Null hypothesis (H₀)**: The mean HIGHPOINT is the same across groups.

-   **Alternative hypothesis (H₁)**: The mean HIGHPOINT differs between groups.

We used a **Welch two-sample t-test**, allowing for unequal variances. The test provides statistical evidence on whether the use of oxygen is associated with higher or lower summit outcomes.

```{r}
t.test(HIGHPOINT ~ O2USED, data = model_data)
```

-   **Result**: The difference is highly significant (*p* \< 2.2e-16).

-   Expeditions that used oxygen reached a **much higher average altitude**\
    (mean = **8465.73 m**)\
    compared to those that did not (mean = **6439.17 m**).

-   The 95% confidence interval for the difference in means is approximately **[1969.9, 2083.2] meters**

This suggests that oxygen use is **strongly associated with better expedition performance**, likely due to its physiological benefits at extreme altitudes.

```{r}
aov_result <- aov(HIGHPOINT ~ SEASON, data = model_data)
summary(aov_result)
```

#### Does season matter?

To investigate whether the time of year influences expedition outcomes, we performed a one-way ANOVA on `HIGHPOINT` across different seasons.

-   **Result:** The effect of `SEASON` on `HIGHPOINT` is highly significant (*p* \< 2.2e–16)

-   This suggests that the season during which an expedition takes place has a measurable influence on how high expeditions typically go.

-   This result supports the inclusion of `SEASON` as a meaningful predictor in the regression models.

```{r}
oxygen_ci <- model_data %>%
  specify(HIGHPOINT ~ O2USED) %>%
  generate(reps = 1000, type = "bootstrap") %>%
  calculate(stat = "diff in means", order = c(TRUE, FALSE))

get_confidence_interval(oxygen_ci, level = 0.95, type = "percentile")

visualize(oxygen_ci) +
  geom_vline(xintercept = mean(oxygen_ci$stat), linetype = "dashed")
```

To further validate the t-test results regarding oxygen use, we ran a **bootstrap simulation** of the difference in means between expeditions that used and did not use supplementary oxygen.

The resulting distribution confirmed a **consistently positive difference**, with the entire 95% confidence interval above zero.\
This reinforces our previous conclusion that expeditions using oxygen tend to reach **significantly higher altitudes**.

```{r}
# Boxplot by HOST
ggplot(model_data, aes(x = factor(HOST_FACTOR), y = HIGHPOINT)) +
  geom_boxplot()
```

We examined the role of the expedition host country (`HOST_FACTOR`) on performance, measured by `HIGHPOINT`. The boxplot below shows the distribution of altitudes reached for expeditions hosted by China, India, Nepal, or Unknown regions.

-   Expeditions hosted by **China** and **Nepal** generally reach higher altitudes, with **China** showing the highest median `HIGHPOINT`.

-   **India** displays a slightly lower median and a narrower range of outcomes.

-   Expeditions with **Unknown** host information show a more moderate distribution.

These differences suggest that the host country may reflect not only geographic access but also logistical, regulatory, or seasonal factors that influence expedition success. Based on this, `HOST_FACTOR` is a meaningful categorical variable to include in the regression models.

```{r}
model_data %>%
  dplyr::select(where(is.numeric)) %>%
  cor(use = "complete.obs") %>%
  ggcorrplot::ggcorrplot(lab = TRUE)

```

correlation matrix was computed to assess linear relationships between all numeric predictors and the target variable `HIGHPOINT`.

Overall, these findings motivate the use of a standard **multiple linear regression model** to assess direction and significance of coefficients, and a **LASSO regression** to deal with multicollinearity and perform feature selection in a more robust way.

**Multiple linear regression and LASSO regression model**

We do a split of the dataset

```{r}
set.seed(42)
split <- initial_split(model_data, prop = 0.8)
train_data <- training(split)
test_data  <- testing(split)
```

To better understand which features contribute to higher performance, we fit two models:

1.  A **multiple linear regression** to interpret coefficient effects.

2.  A **LASSO regression** to evaluate feature importance and perform variable selection.

```{r}
lm_model <- linear_reg() %>%
  set_engine("lm")

lm_recipe <- recipe(HIGHPOINT ~ ., data = train_data) %>%
  step_dummy(all_nominal(), one_hot = TRUE)

lm_wf <- workflow() %>%
  add_model(lm_model) %>%
  add_recipe(lm_recipe)

lm_fit <- lm_wf %>% fit(data = train_data)

# Predict + metrics
pred_lm <- predict(lm_fit, test_data) %>% bind_cols(test_data)
metrics(pred_lm, truth = HIGHPOINT, estimate = .pred)
```

Model performance on the test set:

-   **RMSE**: 1370.5 meters

-   **R²**: 0.5

-   **MAE**: 762.7 meters

This indicates that the model explains around **52% of the variance** in the maximum altitude reached, which is reasonably good given the complexity and variability of real-world expeditions.

and then we do a residual check

```{r}
lm_fit %>% 
  extract_fit_engine() %>%
  check_model()
```

**Normality of residuals** is moderately violated, with noticeable skew in the Q-Q plot.

Now we can implement a LASSO regression

```{r}
exped_full <- exped_full |>
  mutate(HIGHPOINT_SCALED = scale(HIGHPOINT))
```

```{r}
lasso_model <- linear_reg(penalty = tune(), mixture = 1) %>%
  set_engine("glmnet")
```

```{r}
lasso_recipe <- recipe(HIGHPOINT ~ ., data = train_data) %>%
  # Convert logical to numeric (TRUE/FALSE → 1/0)
  step_mutate(O2USED = as.numeric(O2USED),
            O2CLIMB = as.numeric(O2CLIMB),
            COMRTE = as.numeric(COMRTE),
            STDRTE = as.numeric(STDRTE)) %>%
  # One-hot encoding for factors
  step_dummy(all_nominal_predictors(), one_hot = TRUE) %>%
  # Remove constant columns (if any)
  step_zv(all_predictors())
```

```{r}
lasso_wf <- workflow() %>%
  add_model(lasso_model) %>%
  add_recipe(lasso_recipe)
```

```{r}
lasso_res <- tune_grid(
  lasso_wf,
  resamples = vfold_cv(train_data, v = 5),
  grid = 20
)
```

```{r}
best_lasso <- select_best(lasso_res, metric="rmse")
```

```{r}
final_lasso <- finalize_workflow(lasso_wf, best_lasso) %>%
  fit(data = train_data)
```

```{r}
pred_lasso <- predict(final_lasso, test_data) %>%
  bind_cols(test_data)

metrics(pred_lasso, truth = HIGHPOINT, estimate = .pred)
```

To improve model interpretability and reduce the risk of overfitting, we implemented a **LASSO regression**. This method applies L1 regularization, which can shrink less important coefficients to zero, effectively performing variable selection.

We scaled the target variable (`HIGHPOINT`) and performed 5-fold cross-validation to tune the penalty parameter (`λ`), selecting the value that minimized RMSE.

Model performance on the test set:

-   **RMSE**: 1371.1 meters

-   **R²**: 0.519

-   **MAE**: 761.8 meters

```{r}
final_lasso %>%
  pull_workflow_fit() %>%
  tidy() %>%
  filter(estimate != 0)
```

The LASSO regression retained several key predictors, including oxygen use, number of camps, and summit team size.

The size and direction of the coefficients provide insight into each variable's impact on expedition success (measured by HIGHPOINT).

For example, **using supplementary oxygen** and **setting up more camps** are strongly associated with **reaching higher altitudes**, while **larger teams** may slightly reduce performance.

Since both models are linear and the dataset contains a manageable number of informative predictors, their performance is almost identical. LASSO did not substantially outperform the standard linear regression, but it confirms the same key variables while providing a slightly more regularized model

### Classification

#### 3. Are autumn expeditions more likely to fail than spring expeditions?

This analysis explores whether expeditions undertaken during the summer are less successful than those in the spring. Seasonality is a key factor in Himalayan mountaineering.

This is a classification problem: - Response variable: `SUCCESS1` (TRUE/FALSE) - Predictor: `SEASON_FACTOR` (filtered to Spring and Autumn)

Understanding this relationship can help climbers plan their expeditions for better outcomes.

```{r}
data_season <- exped_full %>%
  filter(SEASON_FACTOR %in% c("Spring", "Autumn")) %>%
  select(SUCCESS1, SEASON_FACTOR, TOTMEMBERS, ROPE, O2USED) %>%
  drop_na()

table(data_season$SUCCESS1)
prop.table(table(data_season$SUCCESS1))
```

```{r}
data_season <- data_season %>%
  mutate(SUCCESS1 = as.factor(SUCCESS1))
```

```{r}
# Success rate by season
data_season %>%
  group_by(SEASON_FACTOR) %>%
  summarise(success_rate = mean(SUCCESS1), n = n())

# Bar chart
ggplot(data_season, aes(x = SEASON_FACTOR, fill = SUCCESS1)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Success Rate by Season", y = "Proportion")
```

The bar plot above illustrates the proportion of successful and unsuccessful expeditions for the **Spring** and **Autumn** seasons. While success appears slightly more frequent in **Spring**, the difference is relatively subtle.

This visualization motivates the need for a formal statistical test (e.g., **logistic regression**) to determine whether the observed difference is statistically significant.

-   **Null Hypothesis (H0)**: The probability of success is the same in both seasons.
-   **Alternative Hypothesis (H1)**: The probability of success differs between summer and spring.

```{r}
infer_test <- data_season %>%
  specify(SUCCESS1 ~ SEASON_FACTOR, success = "TRUE") %>%
  generate(reps = 1000, type = "bootstrap") %>%
  calculate(stat = "diff in props", order = c("Autumn", "Spring"))

get_confidence_interval(infer_test, level = 0.95, type = "percentile")

visualize(infer_test) +
  geom_vline(xintercept = mean(infer_test$stat), linetype = "dashed") +
  labs(title = "Bootstrap CI: Spring vs Autumn Success Rate")
```

We use **bootstrap resampling** to estimate the 95% confidence interval for the **difference in success proportions** between autumn and spring expeditions.

The statistic of interest is the difference in porportions between Autumn and Spring.

From the plot, the entire confidence interval lies **below zero**, suggesting that the difference is statistically significant at the 5% level. This reinforces the conclusion that **autumn expeditions are less likely to succeed** than spring ones.

We now split the dataset and prepare it for the logistic regression

```{r}
set.seed(123)
split_season <- initial_split(data_season, prop = 0.8)
train_season <- training(split_season)
test_season  <- testing(split_season)
```

```{r}
rec_season <- recipe(SUCCESS1 ~ ., data = train_season) %>%
  step_dummy(all_nominal_predictors(), one_hot = TRUE)
```

**Logistic Regression**

```{r}

log_model <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

wf_log <- workflow() %>%
  add_model(log_model) %>%
  add_recipe(rec_season)

fit_log <- fit(wf_log, data = train_season)
```

Evaluation

```{r}
log_metrics <- log_preds %>%
  metrics(truth = SUCCESS1, estimate = .pred_class) %>%
  bind_rows(precision(log_preds, truth = SUCCESS1, estimate = .pred_class)) %>%
  bind_rows(recall(log_preds, truth = SUCCESS1, estimate = .pred_class)) %>%
  bind_rows(f_meas(log_preds, truth = SUCCESS1, estimate = .pred_class))

log_metrics
```

```{r}
log_metrics <- log_preds %>%
  metrics(truth = SUCCESS1, estimate = .pred_class) %>%
  bind_rows(precision(log_preds, truth = SUCCESS1, estimate = .pred_class)) %>%
  bind_rows(recall(log_preds, truth = SUCCESS1, estimate = .pred_class)) %>%
  bind_rows(f_meas(log_preds, truth = SUCCESS1, estimate = .pred_class))

log_metrics
```

We fit a **logistic regression** to predict the probability of expedition success (`SUCCESS1`) based on predictors such as **season**, **team size**, **oxygen use**, and **rope usage**, focusing only on **spring** and **autumn** expeditions.

We evaluate model performance on the test set using two metrics:

-   **Accuracy**: 0.699 — This means the model correctly predicts success or failure about 70% of the time

These results suggest that season, along with a few other predictors, offers **some discriminatory power** for predicting expedition success, but there’s still **room for improvement**, perhaps using non-linear models or additional variables.

```{r}
# Confusion matrix
conf_mat(log_preds, truth = SUCCESS1, estimate = .pred_class) %>%
  autoplot(type = "heatmap") +
  labs(title = "Confusion Matrix - Logistic Regression")
```

```{r}
# ROC Curve
log_preds %>%
  roc_curve(truth = SUCCESS1, .pred_TRUE) %>%
  autoplot() +
  ggtitle("ROC Curve - Logistic Regression")

```

The **confusion matrix** reveals a reasonable balance between correctly classified successes and failures, although there are some false positives and false negatives.

The **ROC curve** shows moderate discriminatory power. While the model is better than random guessing, its predictive performance could be improved with additional features or more complex models.

```{r}
# Naive Bayes (Optional Comparison)
library("discrim")
nb_model <- naive_Bayes() %>%
  set_engine("klaR") %>%
  set_mode("classification")

wf_nb <- workflow() %>%
  add_model(nb_model) %>%
  add_recipe(rec_season)

fit_nb <- fit(wf_nb, data = train_season)

nb_preds <- predict(fit_nb, test_season, type = "prob") %>%
  bind_cols(predict(fit_nb, test_season)) %>%
  bind_cols(test_season)
```

```{r}
# Evaluation
metrics(nb_preds, truth = SUCCESS1, estimate = .pred_class)
conf_mat(nb_preds, truth = SUCCESS1, estimate = .pred_class)
roc_auc(nb_preds, truth = SUCCESS1, .pred_TRUE)
```

```{r}
# Confusion matrix
conf_mat(nb_preds, truth = SUCCESS1, estimate = .pred_class) %>%
  autoplot(type = "heatmap") +
  labs(title = "Confusion Matrix - Naive Bayes")
```

As an additional comparison, we applied a Naive Bayes classifier to predict expedition success based on the season. This probabilistic model assumes feature independence and is often used as a simple yet effective baseline in classification tasks.

The confusion matrix shows a higher number of false positives and false negatives compared to logistic regression, indicating slightly lower predictive performance. However, Naive Bayes still captures the overall trend that **spring expeditions are more likely to succeed than autumn ones**.

We tested whether expedition success differs between spring and autumn.

-   **Bootstrap simulation** confirmed a small but statistically significant difference in success rate, with spring expeditions being slightly more successful.

-   **Logistic regression** achieved \~70% accuracy, suggesting season alone carries moderate predictive power.

-   **Naive Bayes**, though slightly less accurate (65%), supports the same conclusion.

**Takeaway**: While seasonality matters, it's not the sole driver of success. Other factors like team size, oxygen use, and route likely play substantial roles.

#### 4. Does the use of supplementary oxygen increase the probability of success?

We aim to determine whether using supplementary oxygen (`O2USED`) increases the probability that an expedition successfully reaches the summit (`SUCCESS1`).

This is a binary classification task: - Response variable: `SUCCESS1` (TRUE/FALSE) - Predictor: `O2USED` (TRUE/FALSE)

We'll explore the data, perform hypothesis testing, fit a logistic regression model, and evaluate its predictive performance. We also compare results with a random forest classifier.

```{r}
# Subset and clean
data_cls <- exped_full %>%
  dplyr::select(SUCCESS1, O2USED, TOTMEMBERS, SMTMEMBERS, O2CLIMB, SEASON_FACTOR, HOST_FACTOR, CAMPS, ROPE, COMRTE, STDRTE) %>%
  drop_na()
```

```{r}
data_cls %>%
  group_by(O2USED) %>%
  summarise(success_rate = mean(SUCCESS1 == "TRUE"), count = n())
```

```{r}
data_cls <- data_cls %>%
  mutate(SUCCESS1 = as.factor(SUCCESS1))
```

```{r}
table(data_cls$O2USED, data_cls$SUCCESS1)
prop.table(table(data_cls$O2USED, data_cls$SUCCESS1), margin = 1)
```

```{r}
ggplot(data_cls, aes(x = O2USED, fill = SUCCESS1)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Success rate by Oxygen Usage", y = "Proportion")
```

The dataset is imbalanced: expeditions using supplementary oxygen (O2USED = TRUE) have a significantly higher success rate (84%) compared to those not using oxygen (45%). This imbalance inflates classification accuracy and motivates the use of additional evaluation metrics.

-   Expeditions that **used supplementary oxygen** (`O2USED = TRUE`) show a **substantially higher proportion of success** compared to those that did not.

To statistically evaluate the effect of supplementary oxygen on success rate, we formulate the following hypotheses:

-   **Null Hypothesis (H₀):** The probability of success is the same for expeditions that used oxygen and those that didn’t.
-   **Alternative Hypothesis (H₁):** The probability of success is different between groups.

We use bootstrap resampling to generate the sampling distribution of the difference in proportions and construct a confidence interval using the `infer` package.

```{r}
# Bootstrap difference in proportions
oxygen_boot <- data_cls %>%
  specify(SUCCESS1 ~ O2USED, success = "TRUE") %>%
  generate(reps = 1000, type = "bootstrap") %>%
  calculate(stat = "diff in props", order = c(TRUE, FALSE))

get_confidence_interval(oxygen_boot, level = 0.95, type = "percentile")

visualize(oxygen_boot)
```

Expeditions using supplementary oxygen have a **substantially and significantly higher probability** of reaching the summit. The evidence supports rejecting the null hypothesis of equal success rates.

Now again let's split the dataset and prepare it for the logistic regression

```{r}
set.seed(123)
split_cls <- initial_split(data_cls, prop = 0.8)
train_cls <- training(split_cls)
test_cls  <- testing(split_cls)
```

```{r}
cls_recipe <- recipe(SUCCESS1 ~ ., data = train_cls) %>%
step_dummy(all_nominal_predictors(), one_hot = TRUE)

```

**Logistic Regression Model**

```{r}
log_model <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

log_wf <- workflow() %>%
  add_model(log_model) %>%
  add_recipe(cls_recipe)

log_fit <- fit(log_wf, data = train_cls)
```

**Evaluate Logistic Regression**

```{r}
log_metrics <- log_preds %>%
  metrics(truth = SUCCESS1, estimate = .pred_class) %>%
  bind_rows(precision(log_preds, truth = SUCCESS1, estimate = .pred_class)) %>%
  bind_rows(recall(log_preds, truth = SUCCESS1, estimate = .pred_class)) %>%
  bind_rows(f_meas(log_preds, truth = SUCCESS1, estimate = .pred_class))

log_metrics
```

#### Model Evaluation

-   **Accuracy**: 0.99
-   Precision: 0.98
-   Recall: 0.99

While the logistic regression model achieves extremely high accuracy (98.8%) and F1-score (98.5%), this performance is partially driven by class imbalance: expeditions using supplementary oxygen are much more likely to succeed. The high recall (99.2%) and precision (97.9%) suggest that the model rarely misses true summit successes and seldom overpredicts success.

However, given the data imbalance, these results should be interpreted cautiously. The model is effective, but its strong performance may be largely attributed to the dominant pattern in the data (oxygen use ≈ success). Additional validation with more balanced data or alternative modeling strategies (e.g., downsampling, SMOTE) could strengthen these findings.

```{r}
# Metrics and confusion matrix for logistic regression
log_preds <- predict(log_fit, test_cls, type = "prob") %>%
  bind_cols(predict(log_fit, test_cls)) %>%
  bind_cols(test_cls)

# Accuracy and other metrics
metrics(log_preds, truth = SUCCESS1, estimate = .pred_class)

# Confusion matrix
conf_mat(log_preds, truth = SUCCESS1, estimate = .pred_class) %>%
  autoplot(type = "heatmap") +
  labs(title = "Confusion Matrix – Logistic Regression")
```

```{r}
log_preds %>%
  roc_curve(truth = SUCCESS1, .pred_TRUE) %>%
  autoplot() +
  ggtitle("ROC Curve – Logistic Regression")
```

Now let's implement a Random Forest model for the comparison

```{r}
rf_model <- rand_forest(trees = 500) %>%
  set_engine("ranger") %>%
  set_mode("classification")

rf_wf <- workflow() %>%
  add_model(rf_model) %>%
  add_recipe(cls_recipe)

rf_fit <- fit(rf_wf, data = train_cls)

rf_preds <- predict(rf_fit, test_cls, type = "prob") %>%
  bind_cols(predict(rf_fit, test_cls)) %>%
  bind_cols(test_cls)

metrics(rf_preds, truth = SUCCESS1, estimate = .pred_class)
roc_auc(rf_preds, truth = SUCCESS1, .pred_TRUE)
```

Both the logistic regression and the random forest model showed excellent performance in predicting expedition success, with very high accuracy and agreement. The results clearly suggest that using supplementary oxygen significantly increases the probability of a successful climb.

#### 5. **Are expeditions hosted by Nepal or China more likely to succeed?**

We aim to assess whether the country leading an expedition (Nepal or China) influences the probability of successfully reaching the summit (`SUCCESS1`).

This is a binary classification task where: - The response variable is `SUCCESS1` (TRUE/FALSE) - The main predictor is `HOST_FACTOR` (Nepal or China)

We will explore class balance, perform hypothesis testing, fit a logistic regression model, and evaluate it using AUC, ROC, and a confusion matrix.

```{r}
# Check target distribution
table(exped_full$SUCCESS1)
prop.table(table(exped_full$SUCCESS1))
```

We begin by checking the distribution of the target variable, `SUCCESS1`. The success rate is relatively balanced, with approximately 58% of expeditions being successful. This indicates we can proceed with standard classification metrics without needing to address severe class imbalance.

Next, we subset the data to focus only on expeditions hosted by **Nepal** or **China**, as these are the two countries we are comparing. We also remove missing values and drop unused factor levels to ensure a clean dataset for modeling.

```{r}
# Subset relevant variables
data_cls <- exped_full %>%
  filter(HOST_FACTOR %in% c("Nepal", "China")) %>%
  dplyr::select(SUCCESS1, HOST_FACTOR, TOTMEMBERS, SMTMEMBERS, O2USED, CAMPS, ROPE, COMRTE, O2CLIMB, SEASON_FACTOR, STDRTE) %>% 
  drop_na()
```

```{r}
table(data_cls$HOST_FACTOR)
data_cls <- droplevels(data_cls)
```

```{r}
data_cls <- data_cls %>%
  mutate(SUCCESS1 = as.factor(SUCCESS1))
```

```{r}
data_cls %>%
  group_by(HOST_FACTOR) %>%
  summarise(success_rate = mean(SUCCESS1), n = n())

ggplot(data_cls, aes(x = HOST_FACTOR, fill = SUCCESS1)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Success rate by Host Country", y = "Proportion")
```

We test if success rates differ significantly between expeditions led by Nepal or China.

-   **Null Hypothesis (H₀)**: All countries have the same success rate.
-   **Alternative Hypothesis (H₁)**: At least one country differs in success rate.

```{r}
chisq.test(table(data_cls$HOST_FACTOR, data_cls$SUCCESS1))
```

A **Chi-squared test** yielded a significant p-value (p = 0.00127), suggesting the host country is associated with differing success rates.

```{r}
data_host_pair <- data_cls %>%
  filter(HOST_FACTOR %in% c("Nepal", "China")) %>%
  droplevels()

inference_host <- data_host_pair %>%
  specify(SUCCESS1 ~ HOST_FACTOR, success = "TRUE") %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "diff in props", order = c("Nepal", "China"))

# Visualize + observed
obs_diff <- data_host_pair %>%
  group_by(HOST_FACTOR) %>%
  summarise(prop_success = mean(SUCCESS1 == "TRUE"))

visualize(inference_host) +
  geom_vline(xintercept = diff(obs_diff$prop_success), linetype = "dashed")
```

The dashed vertical line represents the **observed difference in success rates** between expeditions led by Nepal and China. Since this observed value lies **far in the tail** of the simulated null distribution, it suggests that such a difference is **unlikely to occur by chance**.

This supports the conclusion that **expedition success is significantly associated with host country**, with China-led expeditions showing a higher success rate.

But let's implement all the pipeline for the logistic regression model to see if it's true:

```{r}
set.seed(123)
split_cls <- initial_split(data_cls, prop = 0.8)
train_cls <- training(split_cls)
test_cls  <- testing(split_cls)
```

```{r}
cls_recipe <- recipe(SUCCESS1 ~ ., data = train_cls) %>%
  step_dummy(all_nominal_predictors(), one_hot = TRUE)
```

**Logistic Regression Model**

```{r}
log_model <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

log_wf <- workflow() %>%
  add_model(log_model) %>%
  add_recipe(cls_recipe)

log_fit <- fit(log_wf, data = train_cls)
```

p-value of logistic regression

```{r}
log_fit %>%
  pull_workflow_fit() %>%
  tidy() %>%
  filter(term != "(Intercept)") %>%
  arrange(p.value)
```

P-values provide insight into which predictors significantly contribute to the model. A p-value \< 0.05 indicates a statistically significant effect.

```{r}
exped_full %>%
  filter(HOST_FACTOR %in% c("Nepal", "China")) %>%
  group_by(HOST_FACTOR) %>%
  summarise(mean_peak_height = mean(HEIGHTM, na.rm = TRUE))
```

Interestingly, expeditions hosted by China target significantly higher peaks on average (mean: 8440 m) compared to those hosted by Nepal (mean: 7669 m). Despite this, China-led expeditions still show a higher unadjusted success rate. This further suggests that differences in success are not driven by easier objectives, but rather by operational or strategic differences — such as oxygen use, team size, or route choice — which are better accounted for in the multivariable logistic regression.

**Evaluation**

```{r}
log_preds <- predict(log_fit, test_cls, type = "prob") %>%
  bind_cols(predict(log_fit, test_cls)) %>%
  bind_cols(test_cls)
```

```{r}
metrics(log_preds, truth = SUCCESS1, estimate = .pred_class)
```

```{r}
conf_mat(log_preds, truth = SUCCESS1, estimate = .pred_class) %>%
  autoplot(type = "heatmap") +
  labs(title = "Confusion Matrix - Logistic Regression")
```

```{r}
log_preds %>%
  roc_curve(truth = SUCCESS1, .pred_TRUE) %>%
  autoplot() +
  ggtitle("ROC Curve - Logistic Regression")

roc_auc(log_preds, truth = SUCCESS1, .pred_TRUE)
```

We then fit a logistic regression model including the host country and additional predictors. While overall model performance was strong (accuracy ≈ 0.99, ROC AUC near 1), the host country itself was **not** a statistically significant predictor (p \> 0.05) once other variables were included.

**Takeaway**: While the host country (Nepal vs. China) appears to influence success rates in isolation, this association likely reflects differences in other expedition characteristics. Once key variables like oxygen use, team size, and route difficulty are controlled for, the host country itself no longer significantly predicts success. Therefore, country affiliation is not a primary determinant of outcome — rather, it proxies for other operational differences.

## Conclusion

This project explored key factors influencing expedition success in the Himalayas, using a comprehensive end-to-end data analysis pipeline.

We began with **data cleaning and preprocessing**, addressing missing values, handling high-NA columns, imputing features, removing duplicates, joining datasets, encoding categorical variables, and scaling numerical features. This careful preparation ensured clean, usable data for downstream analysis.

We then investigated five main research questions combining statistical tests and machine learning models:

1.  **Oxygen use** is associated with significantly higher altitudes reached, and groups using supplementary oxygen tend to summit higher.

2.  Both **multiple linear regression** and **LASSO** identified similar predictors of expedition altitude, showing consistent results with comparable performance, likely due to linearity and manageable multicollinearity.

3.  **Seasonality** affects expedition success: expeditions in **spring** had slightly better outcomes than in **autumn**, as confirmed by hypothesis testing and classification models.

4.  **Using oxygen** clearly improves the **probability of success**, confirmed through statistical tests, logistic regression, and classification metrics — with models achieving high accuracy and AUC scores

**The hosting country** (Nepal vs. China) initially showed different success rates, but after adjusting for other variables, the host factor was **not statistically significant** in the logistic model — suggesting that other factors, not the country itself, drive the difference.

Overall, statistical testing combined with predictive modeling provided robust, complementary insights. Some effects observed in univariate analysis disappeared in multivariate contexts, highlighting the value of controlling for multiple expedition factors.

# da qua non considerare

#### 2. Which features help expeditions reach the summit?

In this section, we reformulate the original question by shifting the focus from the **maximum altitude reached (`HIGHPOINT`)** to the **remaining distance to the summit** (`SUMMIT_GAP`). Specifically, we define:

$$
SUMMIT_GAP=HEIGHTM−HIGHPOINT
$$

This continuous variable reflects how far an expedition was from reaching the summit. A value of 0 means the team successfully summited, while higher values indicate they turned around earlier. The goal is to identify factors that **minimize** this gap.

Unlike the previous question, here we include multiple predictors to account for the complexity of an expedition. These include:

-   Team size (`TOTMEMBERS`, `SMTMEMBERS`)

-   Hired staff (`TOTHIRED`, `SMTHIRED`)

-   Use of supplementary oxygen (`O2USED`, `O2CLIMB`)

-   Climbing season (`SEASON`)

-   Hosting country (`HOST`)

-   Logistic factors such as camps, ropes, and support metrics (`CAMPS`, `ROPE`, `COMRTE`, `STDRTE`)

```{r}
# Create target variable before modeling
exped_full <- exped_full %>%
  mutate(SUMMIT_GAP = HEIGHTM - HIGHPOINT)

# Filter only complete cases for modeling
model_data <- exped_full %>%
  select(SUMMIT_GAP, O2USED, TOTMEMBERS, SMTMEMBERS, TOTHIRED, SMTHIRED, O2CLIMB,
         SEASON, HOST, CAMPS, ROPE, COMRTE, STDRTE, HOST_FACTOR) %>%
  drop_na()
```

```{r}
ggplot(model_data, aes(x = SUMMIT_GAP)) +
  geom_histogram(bins = 30, fill = "steelblue") +
  labs(title = "Distribution of Summit Gap (HEIGHTM - HIGHPOINT)") +
  theme_minimal()
```

The histogram of SUMMIT_GAP shows a right-skewed distribution: most expeditions fall short of the summit, while relatively few reach the top (gap close to zero).

### Does oxygen use reduce the summit gap?

```{r}
t.test(SUMMIT_GAP ~ O2USED, data = model_data)
```

**Result**: p \< 2.2e-16.
The Welch t-test shows a highly significant difference (p \< 2.2e-16), with expeditions using supplementary oxygen having a much smaller average summit gap. This suggests oxygen helps expeditions get closer to the summit.

### Does season matter?

```{r}
aov_result <- aov(SUMMIT_GAP ~ SEASON, data = model_data)
summary(aov_result)
```

Season has a statistically significant impact on the summit gap (p \< 2.2e-16), justifying its inclusion in the regression models.

### Bootstrap simulation for oxygen use

```{r}
oxygen_ci <- model_data %>%
  specify(SUMMIT_GAP ~ O2USED) %>%
  generate(reps = 1000, type = "bootstrap") %>%
  calculate(stat = "diff in means", order = c(TRUE, FALSE))

get_confidence_interval(oxygen_ci, level = 0.95, type = "percentile")

visualize(oxygen_ci) +
  geom_vline(xintercept = mean(oxygen_ci$stat), linetype = "dashed")
```

The bootstrap confirms that expeditions using oxygen consistently achieve lower summit gaps.

```{r}
ggplot(model_data, aes(x = factor(HOST_FACTOR), y = SUMMIT_GAP)) +
  geom_boxplot()
```

```{r}
model_data %>%
  select(where(is.numeric)) %>%
  cor(use = "complete.obs") %>%
  ggcorrplot::ggcorrplot(lab = TRUE)
```

```{r}
set.seed(42)
split <- initial_split(model_data, prop = 0.8)
train_data <- training(split)
test_data  <- testing(split)
```

```{r}
lm_model <- linear_reg() %>%
  set_engine("lm")

lm_recipe <- recipe(SUMMIT_GAP ~ ., data = train_data) %>%
  step_dummy(all_nominal(), one_hot = TRUE)

lm_wf <- workflow() %>%
  add_model(lm_model) %>%
  add_recipe(lm_recipe)

lm_fit <- lm_wf %>% fit(data = train_data)

pred_lm <- predict(lm_fit, test_data) %>% bind_cols(test_data)
metrics(pred_lm, truth = SUMMIT_GAP, estimate = .pred)
```
