---
title: "R Notebook"
output: html_notebook
---

# Final project Data Projects & Hackathon 2

# Exploring Mountaineering Success in the Himalayas: An Analysis of Recent Expeditions (2000–2025)

Dataset: <https://github.com/rfordatascience/tidytuesday/tree/main/data/2025/2025-01-21>

## Introduction and problem setting

This project is based on the “Himalayan Database”, a detailed collection of all documented expeditions to the Nepalese Himalayas since 1905. In particular, we will analyse data on expeditions from 2010 onwards, in order to understand the factors influencing the success or failure of recent mountaineering missions.

The main objective is to explore what factors influence the outcome of an expedition. Some key questions:

-   Does the use of supplementary oxygen reduce the duration of expeditions?

-   Which features help expeditions reach the summit?

-   

-   Does the use of supplementary oxygen increase the probability of success?

-   Are expeditions led by Nepal, China or India more likely to succeed?

These questions are particularly relevant given the rich structure of the dataset.

Understanding the role of **oxygen usage** can help assess both its practical value and potential risks in high-altitude climbing. Analyzing **altitude** and **duration** relationships provides insight into how physical challenges scale with mountain difficulty. Seasonal variation is a classic factor in expedition planning, and quantifying its impact on success rates can inform safer scheduling. Finally, exploring **leadership origin** may reflect differences in strategy, experience, or logistical capacity across countries, revealing potential cultural or regional patterns in expedition outcomes.

These questions are not only statistically approachable through regression models, but also contextually meaningful for both mountaineering professionals and enthusiasts.

## Import data

```{r}
################################################################################
### Cleaning data from the Himalayan Database ##################################
################################################################################


library(tidyverse)
library(foreign)
library(httr)
library(zip)
library(ggcorrplot)
library(formattable)
library(VIM)

Sys.setlocale("LC_ALL", "C")  # resolve encoding problem

# Step 1: Setup URL 
url <- "https://www.himalayandatabase.com/downloads/Himalayan%20Database.zip"
zip_file <- tempfile(fileext = ".zip")
extracted_dir <- tempfile()

# Step 2: download ZIP and exctract
GET(url, write_disk(zip_file, overwrite = TRUE))
unzip(zip_file, exdir = extracted_dir)

dbf_file_peaks <- file.path(extracted_dir, "Himalayan Database", "HIMDATA", "peaks.DBF")
dbf_file_exped <- file.path(extracted_dir, "Himalayan Database", "HIMDATA", "exped.DBF")

# Step 5: read DBF and ignore warnings (already verified it)
peaks_data <- suppressWarnings(read.dbf(dbf_file_peaks, as.is = TRUE))
exped_data <- suppressWarnings(read.dbf(dbf_file_exped, as.is = TRUE))

# Step 6: convert into Tibble
peaks_temp <- as_tibble(peaks_data)
exped_temp <- as_tibble(exped_data)

unlink(c(zip_file, extracted_dir), recursive = TRUE)
```

```{r}
###_____________________________________________________________________________
### Add columns that recode some variables that are integers into character
### strings that describe their subject in the peaks dataset
###_____________________________________________________________________________

peaks_tidy <- peaks_temp |> 
  dplyr::mutate(HIMAL_FACTOR = case_when(
    HIMAL == 0  ~ "Unclassified",
    HIMAL == 1  ~ "Annapurna",
    HIMAL == 2  ~ "Api/Byas Risi/Guras",
    HIMAL == 3  ~ "Damodar",
    HIMAL == 4  ~ "Dhaulagiri",
    HIMAL == 5  ~ "Ganesh/Shringi",
    HIMAL == 6  ~ "Janak/Ohmi Kangri",
    HIMAL == 7  ~ "Jongsang",
    HIMAL == 8  ~ "Jugal",
    HIMAL == 9  ~ "Kangchenjunga/Simhalila",
    HIMAL == 10 ~ "Kanjiroba",
    HIMAL == 11 ~ "Kanti/Palchung",
    HIMAL == 12 ~ "Khumbu",
    HIMAL == 13 ~ "Langtang",
    HIMAL == 14 ~ "Makalu",
    HIMAL == 15 ~ "Manaslu/Mansiri",
    HIMAL == 16 ~ "Mukut/Mustang",
    HIMAL == 17 ~ "Nalakankar/Chandi/Changla",
    HIMAL == 18 ~ "Peri",
    HIMAL == 19 ~ "Rolwaling",
    HIMAL == 20 ~ "Saipal",
    TRUE ~ "Unknown" # Default case if a value doesn't match
  ),
  .after = HIMAL
  ) |> 
  mutate(REGION_FACTOR = case_when(
    REGION == 0 ~ "Unclassified",
    REGION == 1 ~ "Kangchenjunga-Janak",
    REGION == 2 ~ "Khumbu-Rolwaling-Makalu",
    REGION == 3 ~ "Langtang-Jugal",
    REGION == 4 ~ "Manaslu-Ganesh",
    REGION == 5 ~ "Annapurna-Damodar-Peri",
    REGION == 6 ~ "Dhaulagiri-Mukut",
    REGION == 7 ~ "Kanjiroba-Far West",
    TRUE        ~ "Unknown" # Default case if a value doesn't match
  ),
  .after = REGION
  ) |> 
  mutate(PHOST_FACTOR = case_when(
    PHOST == 0 ~ "Unclassified",
    PHOST == 1 ~ "Nepal only",
    PHOST == 2 ~ "China only",
    PHOST == 3 ~ "India only",
    PHOST == 4 ~ "Nepal & China",
    PHOST == 5 ~ "Nepal & India",
    PHOST == 6 ~ "Nepal, China & India",
    TRUE       ~ "Unknown" # Default case if a value doesn't match
  ),
  .after = PHOST
  ) |> 
  mutate(PSTATUS_FACTOR = case_when(
    PSTATUS == 0 ~ "Unknown",
    PSTATUS == 1 ~ "Unclimbed",
    PSTATUS == 2 ~ "Climbed",
    TRUE         ~ "Invalid" # Default case if a value doesn't match
  ),
  .after = PSTATUS
  ) |> 
  mutate(across(c(HIMAL_FACTOR, PHOST_FACTOR, PSTATUS_FACTOR), ~ factor(.)))

```

```{r}
exped_tidy <- exped_temp |> 
  dplyr::mutate(YEAR = as.numeric(trimws(YEAR))) |> 
  dplyr::filter(YEAR >= 2000) |> 
  dplyr::mutate(SEASON_FACTOR = case_when(
    SEASON == 0 ~ "Unknown",
    SEASON == 1 ~ "Spring",
    SEASON == 2 ~ "Summer",
    SEASON == 3 ~ "Autumn",
    SEASON == 4 ~ "Winter",
    TRUE ~ NA_character_
  ), .after = SEASON) |> 
  dplyr::mutate(HOST_FACTOR = case_when(
    HOST == 0 ~ "Unknown",
    HOST == 1 ~ "Nepal",
    HOST == 2 ~ "China",
    HOST == 3 ~ "India",
    TRUE ~ NA_character_
  ), .after = HOST) |> 
  dplyr::mutate(TERMREASON_FACTOR = case_when(
    TERMREASON == 0 ~ "Unknown",
    TERMREASON == 1 ~ "Success (main peak)",
    TERMREASON == 2 ~ "Success (subpeak, foresummit)",
    TERMREASON == 3 ~ "Success (claimed)",
    TERMREASON == 4 ~ "Bad weather (storms, high winds)",
    TERMREASON == 5 ~ "Bad conditions (deep snow, avalanching, falling ice, or rock)",
    TERMREASON == 6 ~ "Accident (death or serious injury)",
    TERMREASON == 7 ~ "Illness, AMS, exhaustion, or frostbite",
    TERMREASON == 8 ~ "Lack (or loss) of supplies, support or equipment",
    TERMREASON == 9 ~ "Lack of time",
    TERMREASON == 10 ~ "Route technically too difficult, lack of experience, strength, or motivation",
    TERMREASON == 11 ~ "Did not reach base camp",
    TERMREASON == 12 ~ "Did not attempt climb",
    TERMREASON == 13 ~ "Attempt rumored",
    TERMREASON == 14 ~ "Other"
  ), .after = TERMREASON)

```

```{r}
# not clean dataset
dim(exped_temp)
# clean dataset
dim(exped_tidy)
```

```{r}
# clean dataset
dim(peaks_tidy)
```

```{r}
head(exped_tidy)
```

```{r}
head(peaks_tidy)
```

```{r}
ggplot(exped_temp, aes(x = as.numeric(YEAR))) +
  geom_bar() +
  scale_x_continuous(
    breaks = seq(
      from = min(as.numeric(exped_temp$YEAR), na.rm = TRUE),
      to = max(as.numeric(exped_temp$YEAR), na.rm = TRUE),
      by = 20
    )
  ) +
  labs(
    title = "Number of Expeditions per Year",
    x = "Year",
    y = "Number of Expeditions"
  ) +
  theme_minimal()

```

## Data Cleaning and Preprocessing

The dataset we will use at the end of the data cleaning and preprocessing is a **merged version** of the peaks_tidy and exped_tidy datasets, filtered to include only expeditions from the year 2000 onwards. The main steps of cleaning included:

-    A detailed **description of available features**, distinguishing between numeric, categorical, and logical variables.

-   Handling of missing values:

    -   **Features with more than 40% missing values were discarded**.

    -    For the remaining ones, missing values are being addressed through appropriate **data imputation techniques** (to be finalized, e.g., median/mode for numerical/categorical, or domain-specific strategies).

    Planned additions:

    -    Identify and address duplicate entries.

    -   Standardize categorical levels (e.g., for country names, reasons for termination, etc.).

PEAK_TIDY DATASET DESCRIPTION

| Variable | Class | Description |
|---------------|---------------|------------------------------------------|
| PEAKID | integer | Peak ID, the key field (unique identifier) for each record. |
| PKNAME | integer | Peak name. |
| PKNAME2 | integer | Peak name 2. |
| LOCATION | integer | Location of the peak. |
| HEIGHTM | integer | Peak height in meters. |
| HEIGHTF | integer | Peak height in feet. |
| HIMAL | integer | Mountain range identifier. See HIMAL_FACTOR for names. |
| HIMAL_FACTOR | integer | Mountain range name. |
| REGION | integer | Region identifier. |
| REGION_FACTOR | character | Region name. |
| OPEN | logical | Indicates whether the peak is open for expeditions. |
| UNLISTED | logical | Indicates whether the peak is unlisted in official records. |
| TREKKING | logical | Indicates whether the peak is a trekking peak. |
| TREKYEAR | integer | Year the peak was designated as a trekking peak. |
| RESTRICT | logical | Indicates whether the peak is restricted for climbing or access. |
| PHOST | integer | Primary host country identifier for the peak. See PHOST_FACTOR for names. |
| PHOST_FACTOR | character | Primary host country name corresponding to the PHOST identifier. |
| PSTATUS | integer | Climbing status identifier for the peak. See PSTATUS_FACTOR for names. |
| PSTATUS_FACTOR | character | Climbing status description corresponding to the PSTATUS identifier. |
| PEAKMEMO | character | Additional notes or remarks about the peak. |
| PYEAR | integer | Year of the first recorded climbing attempt on the peak. |
| PSEASON | character | Climbing season associated with the first recorded ascent or expedition. |
| PEXPID | integer | Expedition ID associated with the first recorded climb. |
| PSMTDATE | character | Date of the first successful summit. |
| PCOUNTRY | character | Country of origin for the expedition or climbers. |
| PSUMMITERS | integer | Number of climbers who first reached the summit. |
| PSMTNOTE | character | Notes or remarks related to the first successful summit. |
| REFERMEMO | character | Peak chronology references. |
| PHOTOMEMO | character | Peak photo references. |

EXPED_TIDY DATASET DESCRIPTION

| Variable | Class | Description |
|----------------|----------------|-----------------------------------------|
| EXPID | factor | Expedition ID. |
| PEAKID | factor | Peak ID. |
| YEAR | factor | Year of the expedition. |
| SEASON | integer | Season of the expedition as an integer. |
| SEASON_FACTOR | character | Season of the expedition converted to a descriptive name. |
| HOST | integer | Host country of the expedition as an integer. |
| HOST_FACTOR | character | Host country of the expedition converted to a descriptive name. |
| ROUTE1 | factor | Climbing route 1. |
| ROUTE2 | factor | Climbing route 2. |
| ROUTE3 | factor | Climbing route 3. |
| ROUTE4 | factor | Climbing route 4. |
| NATION | factor | Principal nationality. |
| LEADERS | factor | Leadership of the expedition. |
| SPONSOR | factor | Expedition sponsor / name. |
| SUCCESS1 | logical | Success on route 1? |
| SUCCESS2 | logical | Success on route 2? |
| SUCCESS3 | logical | Success on route 3? |
| SUCCESS4 | logical | Success on route 4? |
| ASCENT1 | factor | Ascent numbers for route 1. |
| ASCENT2 | factor | Ascent numbers for route 2. |
| ASCENT3 | factor | Ascent numbers for route 3. |
| ASCENT4 | factor | Ascent numbers for route 4. |
| CLAIMED | logical | Success claimed? |
| DISPUTED | logical | Success disputed? |
| COUNTRIES | factor | Other countries. |
| APPROACH | factor | Approach march. |
| BCDATE | date | Date arrived at base camp. |
| SMTDATE | date | Date reached summit. |
| SMTTIME | factor | Time reached summit. |
| SMTDAYS | integer | Number of days to summit / high-point. |
| TOTDAYS | integer | Total number of days. |
| TERMDATE | date | Date the expedition was terminated. |
| TERMREASON | integer | Reason the expedition was terminated as an integer. |
| TERMREASON_FACTOR | character | Reason the expedition was terminated converted to a descriptive name. |
| TERMNOTE | factor | Termination details. |
| HIGHPOINT | integer | Expedition high-point. |
| TRAVERSE | logical | Did the expedition traverse. |
| SKI | logical | Ski / snowboard descent? |
| PARAPENTE | logical | Parapente descent? |
| CAMPS | integer | Number of high camps above base camp. |
| ROPE | integer | Amount of fixed rope. |
| TOTMEMBERS | integer | Number of members. |
| SMTMEMBERS | integer | Number of members on summit. |
| MDEATHS | integer | Number of member deaths. |
| TOTHIRED | integer | Number of hired personnel (above base camp). |
| SMTHIRED | integer | Number of hired personnel on summit. |
| HDEATHS | integer | Number of hired personnel deaths. |
| NOHIRED | logical | No hired personnel used above base camp. |
| O2USED | logical | Oxygen used? |
| O2NONE | logical | Oxygen not used? |
| O2CLIMB | logical | Oxygen climbing? |
| O2DESCENT | logical | Oxygen descending? |
| O2SLEEP | logical | Oxygen sleeping? |
| O2MEDICAL | logical | Oxygen used medically? |
| O2TAKEN | logical | Oxygen taken, not used? |
| O2UNKWN | logical | Oxygen use unknown? |
| OTHERSMTS | factor | Other summits. |
| CAMPSITES | factor | Campsite details. |
| ROUTEMEMO | factor | Route details. |
| ACCIDENTS | factor | Accidents on the expedition. |
| ACHIEVMENT | factor | Achievements. |
| AGENCY | factor | Trekking agency. |
| COMRTE | logical | Commercial route? |
| STDRTE | logical | 8000m standard route? |
| PRIMRTE | logical | Route info with primary expedition? |
| PRIMMEM | logical | Member info with primary expedition? |
| PRIMREF | logical | Literature info with primary expedition? |
| PRIMID | factor | Primary expedition ID (if any). |
| CHKSUM | integer | Internal consistency check. |

### Handling NaN

At he moment the data contain a lot of NaN, I want to check If exist some columns that are principaly composed of missing values, if exists, we will remove them

```{r}
# 1. Compute % missing per column
miss_pct <- exped_tidy %>% 
  summarise_all(~ mean(is.na(.)) * 100) %>% 
  pivot_longer(everything(), names_to = "variable", values_to = "pct_missing")

# 2. Filter for columns with > 0% missing values
miss_pct_filtered <- miss_pct %>% 
  filter(pct_missing > 0)

# 3. Plot
miss_pct_filtered %>% 
  ggplot(aes(x = pct_missing, y = fct_reorder(variable, pct_missing))) +
  geom_col() +
  labs(
    x = "% missing",
    y = NULL,
    title = "Missing values per variable in exped_tidy (only > 0%)"
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, .05)))

```

```{r}
missing_pct <- round(colSums(is.na(exped_tidy)) / nrow(exped_tidy) * 100, 2)
missing_df <- data.frame(Column = names(missing_pct), MissingPercentage = missing_pct)

# Create a color-scaled table
formattable(
  missing_df[order(-missing_df$MissingPercentage), ],
  list(
    MissingPercentage = color_tile("white", "red")
  )
)
```

Now we are doing the same for the dataset peaks_tidy

```{r}
# 1. Compute % missing per column
miss_pct <- peaks_tidy %>% 
  summarise_all(~ mean(is.na(.)) * 100) %>% 
  pivot_longer(everything(), names_to = "variable", values_to = "pct_missing")

# 2. Filter for columns with > 0% missing values
miss_pct_filtered <- miss_pct %>% 
  filter(pct_missing > 0)

# 3. Plot
miss_pct_filtered %>% 
  ggplot(aes(x = pct_missing, y = fct_reorder(variable, pct_missing))) +
  geom_col() +
  labs(
    x = "% missing",
    y = NULL,
    title = "Missing values per variable in peaks_tidy (only > 0%)"
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, .05)))
```

```{r}
missing_pct <- round(colSums(is.na(peaks_tidy)) / nrow(peaks_tidy) * 100, 2)
missing_df <- data.frame(Column = names(missing_pct), MissingPercentage = missing_pct)

# Create a color-scaled table
formattable(
  missing_df[order(-missing_df$MissingPercentage), ],
  list(
    MissingPercentage = color_tile("white", "red")
  )
)
```

#### 🔴 Very High NA% (over 40%)

These columns have extremely high missing rates and are usually not useful unless they're absolutely critical:

-   `ASCENT3` 100.00%

-   `ASCENT4` 100.00%

-   `ROUTE4` 99.99%

-   `ROUTE3` 99.87%

-   `ASCENT2` 99.77%

-   `ROUTE2` 98.17%

-   `PRIMID` 95.32%

-   `ACHIEVMENT` 93.69%

-   `ASCENT1` 86.54%

-   `OTHERSMTS` 83.09%

-   `TERMNOTE` 62.30%

-   `COUNTRIES` 61.97%

-   `SMTTIME` 50.39%

-   `APPROACH` 45.40%

-   `REFERMEMO` 96.04%

-   `TREKYEAR` 93.96%

-   `PSMTNOTE` 84.17%

-   `PHOTOMEMO` 83.75%

-   `PKNAME2` 46.46%

-   `RESTRICT` 42.71%

We decided to drop all this columns because of the first analysis of the NaN values

```{r}
cols_to_drop_exped <- c(
  "ASCENT3", "ASCENT4", "ROUTE4", "ROUTE3", "ASCENT2", "ROUTE2", "PRIMID", "ACHIEVMENT",
  "ASCENT1", "OTHERSMTS", "TERMNOTE", "COUNTRIES", "SMTTIME", "APPROACH"
)

exped_tidy <- exped_tidy |> dplyr::select(-all_of(cols_to_drop_exped))

cols_to_drop_peaks <- c(
  "REFERMEMO", "TREKYEAR", "PSMTNOTE", "PHOTOMEMO", "PKNAME2", "RESTRICT"
)

peaks_tidy <- peaks_tidy |> dplyr::select(-all_of(cols_to_drop_peaks))

```

### Handling features that aren't informative

After handling columns with excessive missing values, we proceed to remove additional variables that are not informative for modeling or analysis.

These columns fall into one or more of the following categories: -

-   Descriptive text fields (e.g., notes, comments, or memos) that are unique and not analytically useful

-    Redundant variables where a clearer human-readable factor version already exists

-   High-cardinality identifiers (e.g., names or sponsors) that are not relevant to our research questions

This step is crucial to reduce noise, avoid overfitting in machine learning models, and simplify our dataset for exploration and modeling.

```{r}
# Count unique values per column
unique_counts <- sapply(exped_tidy, function(x) length(unique(x)))

# Sort descending
sort(unique_counts, decreasing = TRUE)
```

```{r}
columns_to_drop_peaks <- c(
  "PKNAME2", "HEIGHTF", "HIMAL_FACTOR", "REGION_FACTOR", "OPEN",
  "UNLISTED", "TREKKING", "TREKYEAR", "RESTRICT", "PHOST_FACTOR",
  "PSTATUS_FACTOR", "PEAKMEMO", "PYEAR", "PSEASON", "PEXPID",
  "PSMTDATE", "PCOUNTRY", "PSUMMITERS", "PSMTNOTE", "REFERMEMO",
  "PHOTOMEMO"
)

columns_to_drop_exped <- c(
  "SEASON_FACTOR", "HOST_FACTOR", "SPONSOR", "ASCENT1", "ASCENT2", "ASCENT3", "ASCENT4", "APPROACH", "TERMREASON_FACTOR", "TERMNOTE", "SKI", "PARAPENTE", "O2NONE", ,"CAMPSITES", "ROUTEMEMO","ACHIEVMENT", "PRIMRTE", "PRIMMEM", "PRIMREF", "PRIMID", "CHKSUM", "ACCIDENTS", "AGENCY", "OTHERSMTS", "LEADERS", "NATION", "COUNTRIES"
)

peaks_tidy <- peaks_tidy |> select(-any_of(columns_to_drop_peaks))

exped_tidy <- exped_tidy |> select(-any_of(c(cols_text_exped, columns_to_drop_exped))

```

### Missing values imputation

To ensure our dataset is complete and ready for modeling, we apply appropriate missing value imputation.

Different strategies are used based on variable types:

-    For numerical variables: we use median imputation to avoid distortion by outliers.

-    For categorical variables: we impute with `"Missing"` or the most frequent category (mode). -

-   For logical variables: missing values are assumed to be `FALSE` when safe, or flagged as `"Unknown"` where interpretation is unclear.

This step ensures that we do not lose rows due to NAs during model fitting.

```{r}
# Imputation - exped_tidy
numeric_vars_exped <- names(exped_tidy)[sapply(exped_tidy, is.numeric)]
exped_numeric <- exped_tidy[, numeric_vars_exped]

# KNN Imputation for numeric vars
exped_numeric_imputed <- kNN(exped_numeric, k = 5, imp_var = FALSE)
exped_tidy[numeric_vars_exped] <- exped_numeric_imputed

# Categorical
cat_vars_exped <- names(exped_tidy)[sapply(exped_tidy, function(x) is.character(x) || is.factor(x))]
for (col in cat_vars_exped) {
  exped_tidy[[col]] <- as.character(exped_tidy[[col]])
  exped_tidy[[col]][is.na(exped_tidy[[col]])] <- "Missing"
  exped_tidy[[col]] <- as.factor(exped_tidy[[col]])
}

# Logical
log_vars_exped <- names(exped_tidy)[sapply(exped_tidy, is.logical)]
for (col in log_vars_exped) {
  exped_tidy[[col]][is.na(exped_tidy[[col]])] <- FALSE
}
```

```{r}
# Imputation - peaks_tidy
numeric_vars_peaks <- names(peaks_tidy)[sapply(peaks_tidy, is.numeric)]
peaks_numeric <- peaks_tidy[, numeric_vars_peaks]

# KNN Imputation
peaks_numeric_imputed <- kNN(peaks_numeric, k = 5, imp_var = FALSE)
peaks_tidy[numeric_vars_peaks] <- peaks_numeric_imputed

# Categorical
cat_vars_peaks <- names(peaks_tidy)[sapply(peaks_tidy, function(x) is.character(x) || is.factor(x))]
for (col in cat_vars_peaks) {
  peaks_tidy[[col]] <- as.character(peaks_tidy[[col]])
  peaks_tidy[[col]][is.na(peaks_tidy[[col]])] <- "Missing"
  peaks_tidy[[col]] <- as.factor(peaks_tidy[[col]])
}

# Logical
log_vars_peaks <- names(peaks_tidy)[sapply(peaks_tidy, is.logical)]
for (col in log_vars_peaks) {
  peaks_tidy[[col]][is.na(peaks_tidy[[col]])] <- FALSE
}
```

```{r}
sum(is.na(exped_tidy)) 
sum(is.na(peaks_tidy))   
```

Every imputation where done correctly, but we have still some missing datas. So let's explore it again

```{r}
na_counts <- colSums(is.na(exped_tidy))
na_counts[na_counts > 0]
```

```{r}
library(lubridate)

exped_tidy <- exped_tidy |> mutate(
  DAYS_TO_END = as.numeric(TERMDATE - BCDATE),
  DAY_OF_YEAR_START = yday(BCDATE)
)
```

```{r}
exped_tidy <- exped_tidy |> select(-BCDATE, -SMTDATE, -TERMDATE)
```

```{r}
colSums(is.na(exped_tidy[c("DAYS_TO_END", "DAY_OF_YEAR_START")]))

# Imputation con KNN se necessario

to_impute <- exped_tidy[, c("DAYS_TO_END", "DAY_OF_YEAR_START")]
to_impute_knn <- kNN(to_impute, k = 5, imp_var = FALSE)
exped_tidy[, c("DAYS_TO_END", "DAY_OF_YEAR_START")] <- to_impute_knn
```

```{r}
# Now should be both 0
sum(is.na(exped_tidy)) 
sum(is.na(peaks_tidy))   
```

We transformed date fields into numeric variables to preserve meaningful temporal information while making them usable in modeling. Specifically, we computed:

-   `DAYS_TO_END`: days from base camp arrival to expedition termination

-   `DAY_OF_YEAR_START`: expedition start day of the year

We removed the original date columns and applied KNN imputation (k = 5) on the derived variables to handle remaining missing values.

### Duplicates

```{r}
duplicated_rows_exped <- exped_tidy[duplicated(exped_tidy), ]
duplicated_rows_peaks <- peaks_tidy[duplicated(peaks_tidy), ]
nrow(duplicated_rows_exped)
nrow(duplicated_rows_peaks)
```

### Join the two datasets

```{r}
exped_full <- exped_tidy |> 
  left_join(peaks_tidy, by = "PEAKID")
```

### Encoding

To make our dataset suitable for modeling, we encoded all categorical variables as factors. This allows R's linear and logistic regression functions (`lm()`, `glm()`) to automatically generate appropriate dummy variables internally.

```{r}
# Convert character columns to factor
exped_full <- exped_full |> mutate(
  across(where(is.character), as.factor)
)
```

```{r}
# List categorical variables
cat_vars <- names(exped_full)[sapply(exped_full, is.factor)]

```

### Scaling

We applied standard scaling to all continuous numeric variables, transforming them to have a mean of 0 and a standard deviation of 1. This step is important for models like linear and logistic regression, where variables on very different scales can distort the estimation of coefficients.

We excluded logical and categorical variables from this process.

```{r}
# Identify numeric variables (excluding logical)
numeric_vars <- names(exped_full)[sapply(exped_full, is.numeric)]
# Optionally exclude known binary numeric vars
binary_like <- c("SUCCESS1", "O2USED", "O2NONE", "O2CLIMB", "O2DESCENT", "O2SLEEP", "O2MEDICAL", "O2TAKEN", "O2UNKWN", "NOHIRED")

vars_to_scale <- setdiff(numeric_vars, binary_like)

# Apply standard scaling (mean = 0, sd = 1)
exped_full_scaled <- exped_full
exped_full_scaled[vars_to_scale] <- scale(exped_full_scaled[vars_to_scale])

# Check: mean should be ~0, sd ~1
summary(exped_full_scaled[vars_to_scale])
```

## Data Exploration (EDA)

Currently underway:

-    Summary statistics and distribution plots for **numerical features**.

-   Planned analysis for:

    -    **Categorical features** (e.g., country, season, termination reason)

    -    **Logical features** (e.g., use of oxygen, success/failure).

-    Exploratory visualizations to support modeling choices (e.g., boxplots, bar charts, correlation heatmaps).

Information about the

Overview of the dataset and the features

```{r}
sapply(exped_full, class) |> table()
```

```{r}
sapply(exped_full, class) |> sort()

```

Analysis of the numerical features:

```{r}
numeric_vars <- names(exped_tidy)[sapply(exped_tidy, is.numeric)]
print(numeric_vars)

```

```{r}
for (var in numeric_vars) {
  print(
    ggplot(exped_tidy, aes(x = .data[[var]])) +
      geom_histogram(bins = 30, fill = "steelblue", color = "white") +
      labs(title = paste("Distribution of", var), x = var, y = "Frequency")
  )
}
```

```{r}
cor_matrix <- exped_full |> 
  dplyr::select(all_of(numeric_vars)) |> 
  cor(use = "complete.obs")

library(ggcorrplot)
ggcorrplot::ggcorrplot(cor_matrix, lab = TRUE, lab_size = 3)
```

```{r}
library(GGally)
exped_full_num <- exped_full |> select(TOTDAYS, SMTDAYS, TOTMEMBERS, SMTMEMBERS, HIGHPOINT)
GGally::ggpairs(exped_full_num)
```

```{r}
categorical_vars <- names(exped_full)[sapply(exped_full, function(x) is.character(x) || is.factor(x))]

print(categorical_vars)

```

```{r}
ggplot(exped_full, aes(x = SEASON_FACTOR)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Distribution of expeditions per season", x = "Season", y = "Count")

```

```{r}
# Distribution by host
exped_full |> count(HOST_FACTOR) |> ggplot(aes(x = HOST_FACTOR, y = n)) +
  geom_bar(stat = "identity") + labs(title = "Distribution by host", x = "Host", y = "expedition")
```

```{r}
# Success per season
exped_tidy |> 
 mutate(success = ifelse(SUCCESS1, "Success", "Failure")) |> 
  count(SEASON_FACTOR, success) |> 
 ggplot(aes(x = SEASON_FACTOR, y = n, fill = success)) +
 geom_bar(stat = "identity", position = "dodge") +
 labs(title = "Successes and Failures by Season", x = "Season", y = "Number", fill = "Outcome")
```

```{r}
# Oxygen and success
exped_tidy |> 
 mutate(O2_USED = ifelse(O2USED, "O2 Used", "No O2")) |> 
 mutate(success = ifelse(SUCCESS1, "Success", "Failure")) |> 
  count(O2_USED, success) |> 
  ggplot(aes(x = O2_USED, y = n, fill = success)) +
 geom_bar(stat = "identity", position = "dodge") +
 labs(title = "Success in relation to Oxygen", x = "Oxygen", y = "Number", fill = "Result")

```

```{r}
# Boxplot giorni
ggplot(exped_tidy, aes(y = TOTDAYS)) + 
  geom_boxplot() + 
  labs(title = "Outliers in the total expedition days", y = "tot days")

```

```{r}

# outliers in members
exped_tidy |> filter(TOTMEMBERS > 30) |> select(EXPID, TOTMEMBERS)
```

## Modeling

### Regression

#### Does the use of supplementary oxygen reduce the duration of expeditions?

We aim to understand whether the use of supplementary oxygen (`O2USED`) has a measurable impact on the total duration of Himalayan expeditions (`TOTDAYS`). This is a regression problem, where the response variable is continuous.

The variable `O2USED` is binary (TRUE/FALSE), and our hypothesis is that expeditions using oxygen may be shorter due to improved endurance at altitude. Confirming this relationship can provide insight into the logistical and physiological effects of oxygen use.

We will begin with exploratory inference, then fit a linear regression model.

To statistically assess whether supplementary oxygen affects expedition duration, we test the following hypotheses:

-   **Null hypothesis (H₀):** There is no difference in average expedition duration (`TOTDAYS`) between expeditions that used supplementary oxygen (`O2USED = TRUE`) and those that did not.
-   **Alternative hypothesis (H₁):** Expeditions that used oxygen had a different average duration than those that did not.

We perform a two-sample t-test: \`\`\`{r} t.test(TOTDAYS \~ O2USED, data = exped_tidy)

Boxplot

```{r}
ggplot(exped_tidy, aes(x = O2USED, y = TOTDAYS)) +
  geom_boxplot(fill = "skyblue") +
  labs(title = "Duration of Expeditions by Oxygen Use",
       x = "Supplementary Oxygen Used",
       y = "Total Duration (days)") +
  theme_minimal()

```

\+ summary stats +

```{r}
# Mean duration by oxygen use
exped_tidy |>
  group_by(O2USED) |>
  summarise(
    count = n(),
    mean_days = mean(TOTDAYS),
    sd_days = sd(TOTDAYS)
  )
```

t-test

```{r}
# t-test
t.test(TOTDAYS ~ O2USED, data = exped_tidy)
```

We model the total duration of expeditions (`TOTDAYS`) as a function of oxygen use (`O2USED`), using simple linear regression.

This model helps quantify the expected change in duration when oxygen is used, holding all else constant. Since oxygen use is binary, the coefficient represents the average difference in days between the two groups.

```{r}
# Fit linear regression
lm_oxy <- lm(TOTDAYS ~ O2USED, data = exped_tidy)

# Summary of model
summary(lm_oxy)
```

Residuals

```{r}
# Residual plots
library(performance)
check_model(lm_oxy)
```

The coefficient for `O2USED` represents the estimated average change in duration when oxygen is used.

If the coefficient is negative and statistically significant (p \< 0.05), this suggests that expeditions using supplementary oxygen are shorter on average.

We also assess the model’s R² value to determine how much of the variation in duration is explained by oxygen use alone. Since this is a simple model, we expect a relatively low R², which is acceptable for initial inference.

The residual plots are used to check for any major violations of the linear model assumptions (homoscedasticity, normality of residuals).

## Conclusion
